---
title: "Aphids_DESEQ2"
format: html
editor: visual
---

## **Metatranscriptomics Aphid Project (MAPPED TO ACIDOVORAX)**

```{r}

#Required packages loaded

library(GenomicFeatures)
library(GenomicAlignments)
library(DESeq2)
library(edgeR)
library(ape)
library(ggplot2)
library(rmarkdown)
library(RColorBrewer)
library(knitr)
library(viridis)
library(DESeq2)
library(ggplot2)
library(pheatmap)
library(BiocManager)
library(kableExtra)
library(ggrepel)
library(dplyr)
library(tidyr)


#Loading of GFF A.radicis genome file

gff_file <- '~/Dropbox/Aphid_project_703/RNA_SEQ_A.radicis/256381E_Acidovoraxradicis.gff'
df <- read.gff(gff_file)
ID = rep(NA, nrow(df))
locus_tag = rep(NA, nrow(df))
product = rep(NA, nrow(df))
gene_name = rep(NA, nrow(df))  # Added for gene name
widths <- df$end - df$start + 1

# Extraction of the relevant attributes from the GFF file
for(k in 1:nrow(df)){
  tmp <- unlist(strsplit(df$attributes[k], split = ';'))
  for (i in 1:length(tmp)){
    tmp2 = unlist(strsplit(tmp[i], split = '='))
    if(tmp2[1] == 'ID') ID[k] <- tmp2[2]
    if(tmp2[1] == 'locus_tag') locus_tag[k] <- tmp2[2]
    if(tmp2[1] == 'product') product[k] <- tmp2[2]
    if(tmp2[1] == 'gene') gene_name[k] <- tmp2[2]  # Gene name extraction
  }
}

# Checking if the gene names are missing, if so its replaced with ID or locus_tag
gene_name[is.na(gene_name)] <- ID[is.na(gene_name)]  # Fallback to ID if gene_name is unavaiable

# Fix seqid
df$seqid <- as.vector(df$seqid)
df$seqid[df$seqid == '1'] <- '1 length=5542670 depth=1.00x circular=true'

# Creation dataframe with gene names, strain A.radicis
SB = data.frame(
  strain = rep('Aradicis', nrow(df)),
  seqid = df$seqid,
  type = df$type,
  ID,
  locus_tag,
  product,
  gene_name,  # Add gene_name to the dataframe
  strand = df$strand,
  start = df$start,
  end = df$end,
  widths = widths
)

# GRanges object made to show length and location of gene in the genome, and what strand its on (+ or -)

featuresB <- GRanges(
  seqnames = SB$seqid,
  ranges = IRanges(start = SB$start, end = SB$end),
  strand = SB$strand,
  group_id = SB[, c(1, 3:5)]
)

#Processing of A.radicis mapped BAM files

#Code creates a table of gene expression counts, counting how many mRNA reads map to the     A.radicis genome in each BAM file.

setwd('~/Dropbox/Aphid_project_703/RNA_SEQ_A.radicis/BAM_files_A.radicis')

# Creation of list of BAM files
bamfiles <- list.files(recursive = TRUE, pattern = "*bam$", full.names = TRUE) # * takes all files that contain bam 

# Summmary of mapped reads overlaps
olap <- summarizeOverlaps(featuresB, bamfiles, mode = 'Union')

# Extraction of count matrix
counts <- assay(olap)

# Initialisation of transcripts per million (TPM) matrix
TPMs <- matrix(NA, nrow(counts), ncol(counts))

# 'ind' defined as all row indices
ind <- 1:nrow(counts)

# Computation of TPMs
for(k in 1:ncol(counts)){
  tmp <- counts[ind, k] / (SB$widths[ind] / 1000) 
  #raw counts divided by gene lengths in kilobases for normalisation
  pmF <- sum(tmp) / 1000000 #Divide by a million to get TPM
  TPM <- tmp / pmF
  TPMs[ind, k] <- TPM
}

# Extraction of sample names from BAM filenames
sample_names <- basename(bamfiles)  # Extraction of just the filename
sample_names <- sub("\\.bam$", "", sample_names)  # Removal of the ".bam" extension

# Column names assigned to the count and TPM matrices
colnames(counts) <- sample_names
colnames(TPMs) <- sample_names

# Row names (gene names) included
rownames(counts) <- SB$gene_name  # Use gene names
rownames(TPMs) <- SB$gene_name    # Use gene names

# Wrote to CSV including row and column names
write.csv(counts, 'raw_counts_with_gene_names.csv', row.names = TRUE)
write.csv(TPMs, 'TPMs_with_gene_names.csv', row.names = TRUE)

# Samples ordered by condition (see code comments)

desired_order <- c(
  "RNAseq_data_merged_1-RNA_CGR_01", "RNAseq_data_merged_2-RNA_CGR_02", "RNAseq_data_merged_3-RNA_CGR_03", "RNAseq_data_merged_4-RNA_CGR_04",    # acidovorax day 1 aphids
  "RNAseq_data_merged_13-RNA_CGR_13", "RNAseq_data_merged_14-RNA_CGR_14", "RNAseq_data_merged_15-RNA_CGR_15", "RNAseq_data_merged_16-RNA_CGR_16",  # acidovorax day 7 aphids
  "RNAseq_data_merged_22-RNA_CGR_22", "RNAseq_data_merged_23-RNA_CGR_23", "RNAseq_data_merged_24-RNA_CGR_24", "RNAseq_data_merged_25-RNA_CGR_25",  # acidovorax day 14 aphids
  "RNAseq_data_merged_34-RNA_CGR_34", "RNAseq_data_merged_35-RNA_CGR_35", "RNAseq_data_merged_36-RNA_CGR_36", "RNAseq_data_merged_37-RNA_CGR_37",  # acidovorax day 1 no aphids
  "RNAseq_data_merged_44-RNA_CGR_44", "RNAseq_data_merged_45-RNA_CGR_45", "RNAseq_data_merged_46-RNA_CGR_46",                            # acidovorax day 14 no aphids
  "RNAseq_data_merged_9-RNA_CGR_9", "RNAseq_data_merged_10-RNA_CGR_10", "RNAseq_data_merged_11-RNA_CGR_11", "RNAseq_data_merged_12-RNA_CGR_12",  # No bacteria day 1 aphids
  "RNAseq_data_merged_20-RNA_CGR_20", "RNAseq_data_merged_21-RNA_CGR_21",                                                  # No bacteria day 7 aphids
  "RNAseq_data_merged_30-RNA_CGR_30", "RNAseq_data_merged_31-RNA_CGR_31", "RNAseq_data_merged_32-RNA_CGR_32", "RNAseq_data_merged_33-RNA_CGR_33",  # No bacteria day 14 aphids
  "RNAseq_data_merged_42-RNA_CGR_42", "RNAseq_data_merged_43-RNA_CGR_43",                                                  # No bacteria day 1 no aphids
  "RNAseq_data_merged_50-RNA_CGR_50", "RNAseq_data_merged_51-RNA_CGR_51", "RNAseq_data_merged_52-RNA_CGR_52", "RNAseq_data_merged_53-RNA_CGR_53",   # No bacteria day 14 no aphids
  "RNAseq_data_merged_5-RNA_CGR_5", "RNAseq_data_merged_6-RNA_CGR_6", "RNAseq_data_merged_7-RNA_CGR_7", "RNAseq_data_merged_8-RNA_CGR_8",    # B.Subtilis day 1 aphids
  "RNAseq_data_merged_17-RNA_CGR_17", "RNAseq_data_merged_18-RNA_CGR_18", "RNAseq_data_merged_19-RNA_CGR_19",                            # B.Subtilis day 7 aphids
  "RNAseq_data_merged_26-RNA_CGR_26", "RNAseq_data_merged_27-RNA_CGR_27", "RNAseq_data_merged_28-RNA_CGR_28", "RNAseq_data_merged_29-RNA_CGR_29",  # B.Subtilis day 14 aphids
  "RNAseq_data_merged_38-RNA_CGR_38", "RNAseq_data_merged_39-RNA_CGR_39", "RNAseq_data_merged_40-RNA_CGR_40", "RNAseq_data_merged_41-RNA_CGR_41",  # B.Subtilis day 1 no aphids
  "RNAseq_data_merged_47-RNA_CGR_47", "RNAseq_data_merged_48-RNA_CGR_48", "RNAseq_data_merged_49-RNA_CGR_49"                             # B.Subtilis day 14 no aphids
)
# Your sample names in desired order
# Your original sample names with "-" separator
samples <- c(
  "RNAseq_data_merged_1-RNA_CGR_01", "RNAseq_data_merged_2-RNA_CGR_02", "RNAseq_data_merged_3-RNA_CGR_03", "RNAseq_data_merged_4-RNA_CGR_04",    
  "RNAseq_data_merged_13-RNA_CGR_13", "RNAseq_data_merged_14-RNA_CGR_14", "RNAseq_data_merged_15-RNA_CGR_15", "RNAseq_data_merged_16-RNA_CGR_16",  
  "RNAseq_data_merged_22-RNA_CGR_22", "RNAseq_data_merged_23-RNA_CGR_23", "RNAseq_data_merged_24-RNA_CGR_24", "RNAseq_data_merged_25-RNA_CGR_25",  
  "RNAseq_data_merged_34-RNA_CGR_34", "RNAseq_data_merged_35-RNA_CGR_35", "RNAseq_data_merged_36-RNA_CGR_36", "RNAseq_data_merged_37-RNA_CGR_37",  
  "RNAseq_data_merged_44-RNA_CGR_44", "RNAseq_data_merged_45-RNA_CGR_45", "RNAseq_data_merged_46-RNA_CGR_46",                            
  "RNAseq_data_merged_9-RNA_CGR_9", "RNAseq_data_merged_10-RNA_CGR_10", "RNAseq_data_merged_11-RNA_CGR_11", "RNAseq_data_merged_12-RNA_CGR_12",  
  "RNAseq_data_merged_20-RNA_CGR_20", "RNAseq_data_merged_21-RNA_CGR_21",                                                  
  "RNAseq_data_merged_30-RNA_CGR_30", "RNAseq_data_merged_31-RNA_CGR_31", "RNAseq_data_merged_32-RNA_CGR_32", "RNAseq_data_merged_33-RNA_CGR_33",  
  "RNAseq_data_merged_42-RNA_CGR_42", "RNAseq_data_merged_43-RNA_CGR_43",                                                  
  "RNAseq_data_merged_50-RNA_CGR_50", "RNAseq_data_merged_51-RNA_CGR_51", "RNAseq_data_merged_52-RNA_CGR_52", "RNAseq_data_merged_53-RNA_CGR_53",   
  "RNAseq_data_merged_5-RNA_CGR_5", "RNAseq_data_merged_6-RNA_CGR_6", "RNAseq_data_merged_7-RNA_CGR_7", "RNAseq_data_merged_8-RNA_CGR_8",    
  "RNAseq_data_merged_17-RNA_CGR_17", "RNAseq_data_merged_18-RNA_CGR_18", "RNAseq_data_merged_19-RNA_CGR_19",                            
  "RNAseq_data_merged_26-RNA_CGR_26", "RNAseq_data_merged_27-RNA_CGR_27", "RNAseq_data_merged_28-RNA_CGR_28", "RNAseq_data_merged_29-RNA_CGR_29",  
  "RNAseq_data_merged_38-RNA_CGR_38", "RNAseq_data_merged_39-RNA_CGR_39", "RNAseq_data_merged_40-RNA_CGR_40", "RNAseq_data_merged_41-RNA_CGR_41",  
  "RNAseq_data_merged_47-RNA_CGR_47", "RNAseq_data_merged_48-RNA_CGR_48", "RNAseq_data_merged_49-RNA_CGR_49"
)

# Dash replaced with a dot to match count matrix column names
samples_dot <- gsub("-", ".", samples)

# Matching of conditions labels to samples
conditions <- c(
  rep("Acidovorax_day1_aphids", 4),
  rep("Acidovorax_day7_aphids", 4),
  rep("Acidovorax_day14_aphids", 4),
  rep("Acidovorax_day1_no_aphids", 4),
  rep("Acidovorax_day14_no_aphids", 3),
  rep("NoBacteria_day1_aphids", 4),
  rep("NoBacteria_day7_aphids", 2),
  rep("NoBacteria_day14_aphids", 4),
  rep("NoBacteria_day1_no_aphids", 2),
  rep("NoBacteria_day14_no_aphids", 4),
  rep("Bsubtilis_day1_aphids", 4),
  rep("Bsubtilis_day7_aphids", 3),
  rep("Bsubtilis_day14_aphids", 4),
  rep("Bsubtilis_day1_no_aphids", 4),
  rep("Bsubtilis_day14_no_aphids", 3)
)

# Final metadata dataframe created
metadata <- data.frame(
  sample = samples_dot,
  condition = conditions
)

#Metadata wrote to csv file
write.csv(metadata, "metadata.csv", row.names = FALSE)


# Counts and TPM matrices reordered based on the sample names
counts_reordered <- counts[, desired_order]
TPMs_reordered <- TPMs[, desired_order]

# Column names correctly assigned after reordering
colnames(counts_reordered) <- desired_order
colnames(TPMs_reordered) <- desired_order

# Reordered data wrote to CSVs
write.csv(counts_reordered, 'raw_counts_reordered.csv', row.names = TRUE)
write.csv(TPMs_reordered, 'TPMs_reordered.csv', row.names = TRUE)

# Count data and TPMs loaded
counts <- read.csv('raw_counts_reordered.csv', row.names = 1)
TPMs <- read.csv('TPMs_reordered.csv', row.names = 1)

# raw count data read in
counts = read.csv("raw_counts_reordered.csv", row.names = 1)

#sample columns ensured to match the metadata sample order (53 samples)
count_matrix = as.matrix(counts[,1:53])

# Read in sample metadata
meta = read.csv("metadata.csv", header=T, stringsAsFactors = F)
#make sampleID the rownames
rownames(meta) = meta$Sample

#Set up for the DESeq2 differential expression analysis
### remove outliers here
keep <- which(!meta$sample%in%c('RNAseq_data_merged_50.RNA_CGR_50','RNAseq_data_merged_53.RNA_CGR_53','RNAseq_data_merged_5.RNA_CGR_5', 'RNAseq_data_merged_28.RNA_CGR_28'))
meta <- meta[keep,]
count_matrix <- count_matrix[,keep]

# Setting up of DESeqDataSet
dds <- DESeqDataSetFromMatrix(countData = count_matrix,
                              colData = meta,
                              design = ~condition)

ncol(count_matrix)
nrow(meta)

colnames(count_matrix)[1:5]   # see a preview
rownames(meta)[1:5]           # make sure these align

# Rownames set using metadata
rownames(meta) <- meta$sample  
all(colnames(count_matrix) %in% rownames(meta))

#group as factor set as a factor

dds$condition<- factor(dds$condition, levels = c("Acidovorax_day1_aphids","Acidovorax_day7_aphids", "Acidovorax_day14_aphids", "Acidovorax_day1_no_aphids", "Acidovorax_day14_no_aphids", "NoBacteria_day1_aphids", "NoBacteria_day7_aphids", "NoBacteria_day14_aphids", "NoBacteria_day1_no_aphids", "NoBacteria_day14_no_aphids", "Bsubtilis_day1_aphids", "Bsubtilis_day7_aphids", "Bsubtilis_day14_aphids", "Bsubtilis_day1_no_aphids", "Bsubtilis_day14_no_aphids"))



# any genes that are not expressed i.e. count of 0 in all samples are filtered out
keep <- rowSums(counts(dds)) >= 1

counts
dds <- dds[keep,] #keeps genes that passed filtering

dds <- estimateSizeFactors(dds)#estimation of size factors for each sample for library normalisation
normalized_counts <- counts(dds, normalized=TRUE) #normalisation
dds <- DESeq(dds) # deseq performed

res <- results(dds, independentFiltering=FALSE)



```

Comparing Conditions using deseq2

1.  Day 7 b.subtillis v a.radicis

```{r}

guides(color = guide_legend(override.aes = list(shape = 16)))


# DE results for pairwise compairson
res_day7_bsub_vs_acido <- results(dds, contrast = c("condition", "Bsubtilis_day7_aphids", "Acidovorax_day7_aphids"))

# Removal of NA padj values
res_filtered <- res_day7_bsub_vs_acido[!is.na(res_day7_bsub_vs_acido$padj), ]

#Filtered data converted to dataframe
res_df <- as.data.frame(res_filtered)

# Gene names assigned from original counts (before filtering)
gene_names <- rownames(counts(dds, normalized = FALSE))
res_df$gene <- gene_names[match(rownames(res_df), gene_names)]

# Regulation category added for up or downregulated genes based on log fold change

res_df$regulation <- "Not Sig"
res_df$regulation[res_df$padj < 0.05 & res_df$log2FoldChange > 1] <- "Up"
res_df$regulation[res_df$padj < 0.05 & res_df$log2FoldChange < -1] <- "Down"

top10_genes <- res_df[order(res_df$padj), ][1:10, ]#select top 10 genes by p adj

library(ggrepel)

# Filtering of significant genes
sig_genes <- res_df[res_df$regulation != "Not Sig", ]

# Volcano plot generated with top 10 significant genes labeled

ggplot(res_df, aes(x = log2FoldChange, y = -log10(padj), color = regulation)) +
  geom_point(alpha = 0.6) +
  scale_color_manual(values = c("Up" = "red", "Down" = "blue", "Not Sig" = "grey")) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black") +
  geom_vline(xintercept = 0, linetype = "dotted", color = "black") +
  geom_text_repel(
    data = sig_genes,
    aes(label = gene),
    size = 3,
    box.padding = 0.3,
    point.padding = 0.3,
    max.overlaps = Inf
  ) +
  labs(
    title = "Volcano Plot: Acidovorax vs B.subtilis (Day 7, Aphids)",
    x = "log2 Fold Change",
    y = "-log10 adjusted p-value"
  ) +
  guides(color = guide_legend(override.aes = list(shape = 16))) +
  theme_minimal()

print(top10_genes)
```

2.  A.radicis day 1 v no bacteria (no aphids)

```{r}
# DE results for pairwise compairson

res_day1_acido_vs_nobac <- results(dds, contrast = c("condition", "Acidovorax_day1_no_aphids", "NoBacteria_day1_no_aphids"))

# Removal of NA padj values
res_filtered <- res_day1_acido_vs_nobac[!is.na(res_day1_acido_vs_nobac$padj), ]

# Converted to dataframe & add gene names added
res_df <- as.data.frame(res_filtered)
res_df$gene <- rownames(res_df)

# Regulation category added for up or downregulated genes based on log fold change
res_df$regulation <- "Not Sig"
res_df$regulation[res_df$padj < 0.05 & res_df$log2FoldChange > 1] <- "Up"
res_df$regulation[res_df$padj < 0.05 & res_df$log2FoldChange < -1] <- "Down"

# Filtering of significant genes
sig_genes <- res_df[res_df$regulation != "Not Sig", ]

# Volcano plot generated with top 10 significant genes labeled
ggplot(res_df, aes(x = log2FoldChange, y = -log10(padj), color = regulation)) +
  geom_point(alpha = 0.6) +
  scale_color_manual(values = c("Up" = "red", "Down" = "blue", "Not Sig" = "grey")) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black") +  # Significance threshold line added
  geom_vline(xintercept = 0, linetype = "dotted", color = "black") +             # to show up or down regulated
  geom_text_repel(
    data = sig_genes,
    aes(label = gene),
    size = 3,
    box.padding = 0.3,
    point.padding = 0.3,
    max.overlaps = Inf
  ) +
  labs(
    title = "Volcano Plot: Acidovorax vs No Bacteria
                    (Day 1, No Aphids)",
    x = "log2 Fold Change",
    y = "-log10 adjusted p-value"
  ) +
  guides(color = guide_legend(override.aes = list(shape = 16))) +  # Remove letter symbols from legend
  theme_minimal()







```

Gene table for A.radicis v no bacteria day 1 (no aphids)

```{r}
# Data ordered by adjusted p-value (padj) and top 15 most significant genes selected
top_15_genes <- res_df[order(res_df$padj), ][1:15, ]

# Gene column added from rownames
top_15_genes$Gene <- rownames(top_15_genes)

# Row names reset to prevent them from being displayed as a separate column
rownames(top_15_genes) <- NULL

# Columns reordered: Gene, log2FoldChange, padj
top_15_genes_table <- top_15_genes[, c("Gene", "log2FoldChange", "padj")]

# Values rounded for better presentation
top_15_genes_table$log2FoldChange <- round(top_15_genes_table$log2FoldChange, 2)
top_15_genes_table$padj <- round(top_15_genes_table$padj, 4)


# Table rendered as HTML
top_15_genes_table %>%
  kable("html", col.names = c("Gene", "log2 Fold Change", "Adjusted p-value")) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "responsive"),
    full_width = FALSE,
    position = "center"
  ) %>%
  column_spec(1, bold = TRUE) %>%
  column_spec(2, color = "blue") %>%
  column_spec(3, color = "red") %>%
  add_header_above(c("Top 15 Differentially Expressed Genes" = 3))  # Correct number of columns


# Ensure all of the categories appear even if one has count 0
reg_levels <- c("Up", "Down")
reg_counts <- as.data.frame(table(factor(res_df$regulation, levels = reg_levels)))
colnames(reg_counts) <- c("Regulation", "Count")

# Plotting of bar chart
ggplot(reg_counts, aes(x = Regulation, y = Count, fill = Regulation)) +
  geom_bar(stat = "identity", width = 0.6) +
  scale_fill_manual(values = c("Up" = "red", "Down" = "blue")) +
  labs(
    title = "Differentially Expressed Genes",
    x = "Regulation",
    y = "Number of Genes"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    text = element_text(size = 12)
  )

# Calculation of stats
up_count <- sum(res_df$regulation == "Up")
down_count <- sum(res_df$regulation == "Down")
total_genes <- nrow(res_df)
dysregulated_count <- up_count + down_count
dysregulated_percent <- round((dysregulated_count / total_genes) * 100, 2) # gives as %

# Creation of summary table
reg_summary <- data.frame(
  Metric = c(
    "Upregulated Genes",
    "Downregulated Genes",
    "Total Genes",
    "Percent Dysregulated"
  ),
  Value = c(
    up_count,
    down_count,
    total_genes,
    paste0(dysregulated_percent, "%")
  )
)

reg_summary %>%
  kable("html", col.names = c("Metric", "Value")) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = FALSE,
    position = "center"
  ) %>%
  add_header_above(c("Differential Expression Summary" = 2)) 

#creation of kable extra table for clear presentation of DE summary

```

A.radicis v no bacteria day 14 (no aphids)

```{r}

# DE results selected for Day 14 comparison
res_day14_acido_vs_nobac <- results(dds, contrast = c("condition", "Acidovorax_day14_no_aphids", "NoBacteria_day14_no_aphids"))

# Removal of NA padj values
res_filtered <- res_day14_acido_vs_nobac[!is.na(res_day14_acido_vs_nobac$padj), ]

# Filtered data converted to dataframe
res_df <- as.data.frame(res_filtered)
res_df$gene <- rownames(res_df)

# Regulation category added for up or downregulated genes based on log fold change
res_df$regulation <- "Not Sig"
res_df$regulation[res_df$padj < 0.05 & res_df$log2FoldChange > 1] <- "Up"
res_df$regulation[res_df$padj < 0.05 & res_df$log2FoldChange < -1] <- "Down"

# Filtering of significant genes
sig_genes <- subset(res_df, regulation %in% c("Up", "Down"))

# Volcano plot generated with top 10 significant genes labeled
ggplot(res_df, aes(x = log2FoldChange, y = -log10(padj), color = regulation)) +
  geom_point(alpha = 0.6) +
  scale_color_manual(
    values = c("Up" = "red", "Down" = "blue", "Not Sig" = "grey"),
    labels = c("Up" = "Overexpressed", "Down" = "Underexpressed", "Not Sig" = "Not Significant")
  ) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black") +  # Significance threshold line added
  geom_vline(xintercept = 0, linetype = "dotted", color = "black") +                # shows up or down regulated change line
  geom_text_repel(
    data = sig_genes,
    aes(label = gene),
    size = 3,
    box.padding = 0.3,
    point.padding = 0.3,
    max.overlaps = Inf
  ) +
  labs(
    title = "Volcano Plot: Acidovorax vs No Bacteria 
                   (Day 14, No Aphids)",
    x = "log2 Fold Change",
    y = "-log10 adjusted p-value"
  ) +
  guides(color = guide_legend(title = "Expression")) +
  theme_minimal()

```

Tables for acido v bacteria day 14 no aphids

```{r}
# Data ordered by adjusted p-value (padj) and top 15 most significant genes selected

top_15_genes <- res_df[order(res_df$padj), ][1:15, ]

# Gene column added from rownames
top_15_genes$Gene <- rownames(top_15_genes)

# Row names reset to prevent them from being displayed as a separate column

rownames(top_15_genes) <- NULL

# Columns reordered: Gene, log2FoldChange, padj

top_15_genes_table <- top_15_genes[, c("Gene", "log2FoldChange", "padj")]

# Values rounded for better presentation

top_15_genes_table$log2FoldChange <- round(top_15_genes_table$log2FoldChange, 2)
top_15_genes_table$padj <- round(top_15_genes_table$padj, 4)

# Table rendered as HTML

top_15_genes_table %>%
  kable("html", col.names = c("Gene", "log2 Fold Change", "Adjusted p-value")) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "responsive"),
    full_width = FALSE,
    position = "center"
  ) %>%
  column_spec(1, bold = TRUE) %>%
  column_spec(2, color = "blue") %>%
  column_spec(3, color = "red") %>%
  add_header_above(c("Top 15 Differentially Expressed Genes" = 3))  # Correct number of columns


# Ensure all of the categories appear even if one has count 0

reg_levels <- c("Up", "Down")
reg_counts <- as.data.frame(table(factor(res_df$regulation, levels = reg_levels)))
colnames(reg_counts) <- c("Regulation", "Count")

# Plotting of bar chart
ggplot(reg_counts, aes(x = Regulation, y = Count, fill = Regulation)) +
  geom_bar(stat = "identity", width = 0.6) +
  scale_fill_manual(values = c("Up" = "red", "Down" = "blue")) +
  labs(
    title = "Differentially Expressed Genes",
    x = "Regulation",
    y = "Number of Genes"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    text = element_text(size = 12)
  )

# Calculation of stats
up_count <- sum(res_df$regulation == "Up")
down_count <- sum(res_df$regulation == "Down")
total_genes <- nrow(res_df)
dysregulated_count <- up_count + down_count
dysregulated_percent <- round((dysregulated_count / total_genes) * 100, 2)

# Creation of summary table
reg_summary <- data.frame(
  Metric = c(
    "Upregulated Genes",
    "Downregulated Genes",
    "Total Genes",
    "Percent Dysregulated"
  ),
  Value = c(
    up_count,
    down_count,
    total_genes,
    paste0(dysregulated_percent, "%")
  )
)

# Creation of kable extra table for clear presentation of DE summary

reg_summary %>%
  kable("html", col.names = c("Metric", "Value")) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = FALSE,
    position = "center"
  ) %>%
  add_header_above(c("Differential Expression Summary" = 2))
```

4.  A.radicis v no bacteria day 1 aphids

```{r}
# DE results for pairwise compairson

res_day1_acido_vs_nobac_aphids <- results(dds, contrast = c("condition", "Acidovorax_day1_aphids", "NoBacteria_day1_aphids"))

# Removal of NA padj values

res_filtered <- res_day1_acido_vs_nobac_aphids[!is.na(res_day1_acido_vs_nobac_aphids$padj), ]

# Converted to dataframe & add gene names added

res_df <- as.data.frame(res_filtered)
res_df$gene <- rownames(res_df)

# Regulation category added for up or downregulated genes based on log fold change

res_df$regulation <- "Not Sig"
res_df$regulation[res_df$padj < 0.05 & res_df$log2FoldChange > 1] <- "Up"
res_df$regulation[res_df$padj < 0.05 & res_df$log2FoldChange < -1] <- "Down"

# Filtering of significant genes

top_genes <- res_df[order(res_df$padj), ][1:3, ]


# Volcano plot generated with top 10 significant genes labeled

ggplot(res_df, aes(x = log2FoldChange, y = -log10(padj), color = regulation)) +
  geom_point(alpha = 0.6) +
  scale_color_manual(
    values = c("Up" = "red", "Down" = "blue", "Not Sig" = "grey"),
    labels = c("Up" = "Overexpressed", "Down" = "Underexpressed", "Not Sig" = "Not Significant")
  ) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black") +  # Significance threshold line added
  geom_vline(xintercept = 0, linetype = "dotted", color = "black") +             # shows up or down regulated
  geom_text_repel(
    data = top_genes,
    aes(label = gene),
    size = 3,
    box.padding = 0.3,
    point.padding = 0.3,
    max.overlaps = Inf
  ) +
  labs(
    title = "Volcano Plot: Acidovorax vs No Bacteria 
                 (Day 1, With Aphids)",
    x = "log2 Fold Change",
    y = "-log10 adjusted p-value"
  ) +
  guides(color = guide_legend(title = "Expression")) +  # legend title changed to "Expression"
  theme_minimal()


```

Table of most sig genes: a.radicis v no bacteria day 1 aphids

```{r}
# Data ordered by adjusted p-value (padj) and top 15 most significant genes selected

top_15_genes <- res_df[order(res_df$padj), ][1:45, ]

# Gene column added from rownames

top_15_genes$Gene <- rownames(top_15_genes)

# Row names reset to prevent them from being displayed as a separate column

rownames(top_15_genes) <- NULL

# Columns reordered: Gene, log2FoldChange, padj

top_15_genes_table <- top_15_genes[, c("Gene", "log2FoldChange", "padj")]

# Values rounded for better presentation

top_15_genes_table$log2FoldChange <- round(top_15_genes_table$log2FoldChange, 2)
top_15_genes_table$padj <- round(top_15_genes_table$padj, 4)

# Table rendered as HTML

top_15_genes_table %>%
  kable("html", col.names = c("Gene", "log2 Fold Change", "Adjusted p-value")) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "responsive"),
    full_width = FALSE,
    position = "center"
  ) %>%
  column_spec(1, bold = TRUE) %>%
  column_spec(2, color = "blue") %>%
  column_spec(3, color = "red") %>%
  add_header_above(c("Top 15 Differentially Expressed Genes" = 3))  # Correct number of columns


# Ensure all of the categories appear even if one has count 0

reg_levels <- c("Up", "Down")
reg_counts <- as.data.frame(table(factor(res_df$regulation, levels = reg_levels)))
colnames(reg_counts) <- c("Regulation", "Count")

# Plotting of bar chart

ggplot(reg_counts, aes(x = Regulation, y = Count, fill = Regulation)) +
  geom_bar(stat = "identity", width = 0.6) +
  scale_fill_manual(values = c("Up" = "red", "Down" = "blue")) +
  labs(
    title = "Differentially Expressed Genes",
    x = "Regulation",
    y = "Number of Genes"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    text = element_text(size = 12)
  )

# Calculation of stats

up_count <- sum(res_df$regulation == "Up")
down_count <- sum(res_df$regulation == "Down")
total_genes <- nrow(res_df)
dysregulated_count <- up_count + down_count
dysregulated_percent <- round((dysregulated_count / total_genes) * 100, 2)

# Creation of summary table

reg_summary <- data.frame(
  Metric = c(
    "Upregulated Genes",
    "Downregulated Genes",
    "Total Genes",
    "Percent Dysregulated"
  ),
  Value = c(
    up_count,
    down_count,
    total_genes,
    paste0(dysregulated_percent, "%")
  )
)

#creation of kable extra table for clear presentation of DE summary

reg_summary %>%
  kable("html", col.names = c("Metric", "Value")) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = FALSE,
    position = "center"
  ) %>%
  add_header_above(c("Differential Expression Summary" = 2))


```

5.  A.radicis v no bacteria day 7 (aphids)

```{r}
# DE results for Day 7 aphid comparison
res_day7_acido_vs_nobac_aphids <- results(dds, contrast = c("condition", "Acidovorax_day7_aphids", "NoBacteria_day7_aphids"))

# Removal of NA padj values
res_filtered <- res_day7_acido_vs_nobac_aphids[!is.na(res_day7_acido_vs_nobac_aphids$padj), ]

# Converted to dataframe & add gene names added
res_df <- as.data.frame(res_filtered)
res_df$gene <- rownames(res_df)

# Regulation category added for up or downregulated genes based on log fold change
res_df$regulation <- "Not Sig"
res_df$regulation[res_df$padj < 0.05 & res_df$log2FoldChange > 1] <- "Up"
res_df$regulation[res_df$padj < 0.05 & res_df$log2FoldChange < -1] <- "Down"

# Filtering of significant genes
top_genes <- res_df[order(res_df$padj), ][1:10, ]

# Load ggplot2 and ggrepel
library(ggplot2)
library(ggrepel)

# Volcano plot generated with top 10 significant genes labeled (without label lines)
ggplot(res_df, aes(x = log2FoldChange, y = -log10(padj), color = regulation)) +
  geom_point(alpha = 0.6) +
  scale_color_manual(
    values = c("Up" = "red", "Down" = "blue", "Not Sig" = "grey"),
    labels = c("Up" = "Overexpressed", "Down" = "Underexpressed", "Not Sig" = "Not Significant")
  ) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black") +  # Significance threshold line
  geom_vline(xintercept = 0, linetype = "dotted", color = "black") +             # log2FC = 0 reference
  geom_text_repel(
    data = top_genes,
    aes(label = gene),
    size = 3,
    box.padding = 0.3,
    point.padding = 0.3,
    max.overlaps = Inf,
    segment.color = NA  # Remove connecting lines from labels
  ) +
  labs(
    title = "Volcano Plot: Acidovorax vs No Bacteria (Day 7, With Aphids)",
    x = "log2 Fold Change",
    y = "-log10 adjusted p-value"
  ) +
  guides(color = guide_legend(title = "Expression")) +
  theme_minimal()

# Counting of DEGs by regulation status
deg_counts <- as.data.frame(table(res_df$regulation))
colnames(deg_counts) <- c("Regulation", "Count")

# Preview of counts
print(deg_counts)

```

Table of most sig genes: A.radicis v no bacteria day 7 aphids

```{r}
# Data ordered by adjusted p-value (padj) and top 15 most significant genes selected

top_15_genes <- res_df[order(res_df$padj), ][1:22, ]

# Gene column added from rownames

top_15_genes$Gene <- rownames(top_15_genes)

# Row names reset to prevent them from being displayed as a separate column

rownames(top_15_genes) <- NULL

# Columns reordered: Gene, log2FoldChange, padj

top_15_genes_table <- top_15_genes[, c("Gene", "log2FoldChange", "padj")]

# Values rounded for better presentation

top_15_genes_table$log2FoldChange <- round(top_15_genes_table$log2FoldChange, 2)
top_15_genes_table$padj <- round(top_15_genes_table$padj, 4)

# Table rendered as HTML
top_15_genes_table %>%
  kable("html", col.names = c("Gene", "log2 Fold Change", "Adjusted p-value")) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "responsive"),
    full_width = FALSE,
    position = "center"
  ) %>%
  column_spec(1, bold = TRUE) %>%
  column_spec(2, color = "blue") %>%
  column_spec(3, color = "red") %>%
  add_header_above(c("Top 15 Differentially Expressed Genes" = 3))  # Correct number of columns

# Ensure all of the categories appear even if one has count 0

reg_levels <- c("Up", "Down")
reg_counts <- as.data.frame(table(factor(res_df$regulation, levels = reg_levels)))
colnames(reg_counts) <- c("Regulation", "Count")

# Plotting of bar chart

ggplot(reg_counts, aes(x = Regulation, y = Count, fill = Regulation)) +
  geom_bar(stat = "identity", width = 0.6) +
  scale_fill_manual(values = c("Up" = "red", "Down" = "blue")) +
  labs(
    title = "Differentially Expressed Genes",
    x = "Regulation",
    y = "Number of Genes"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    text = element_text(size = 12)
  )

# Calculation of stats
up_count <- sum(res_df$regulation == "Up")
down_count <- sum(res_df$regulation == "Down")
total_genes <- nrow(res_df)
dysregulated_count <- up_count + down_count
dysregulated_percent <- round((dysregulated_count / total_genes) * 100, 2)

# Creation of summary table
reg_summary <- data.frame(
  Metric = c(
    "Upregulated Genes",
    "Downregulated Genes",
    "Total Genes",
    "Percent Dysregulated"
  ),
  Value = c(
    up_count,
    down_count,
    total_genes,
    paste0(dysregulated_percent, "%")
  )
)

#creation of kable extra table for clear presentation of DE summary

reg_summary %>%
  kable("html", col.names = c("Metric", "Value")) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = FALSE,
    position = "center"
  ) %>%
  add_header_above(c("Differential Expression Summary" = 2))
```

6.  A.radicis v no bacteria day 14 (aphids)

```{r}

# DE results for pairwise compairson

res_day14_acido_vs_nobac_aphids <- results(dds, contrast = c("condition", "Acidovorax_day14_aphids", "NoBacteria_day14_aphids"))
View(as.data.frame(res_day14_acido_vs_nobac_aphids))
# Removal of NA padj values

res_filtered <- res_day14_acido_vs_nobac_aphids[!is.na(res_day14_acido_vs_nobac_aphids$padj), ]

# Converted to dataframe & add gene names added

res_df <- as.data.frame(res_filtered)
res_df$gene <- rownames(res_df)

# Regulation category added for up or downregulated genes based on log fold change

res_df$regulation <- "Not Sig"
res_df$regulation[res_df$padj < 0.05 & res_df$log2FoldChange > 1] <- "Up"
res_df$regulation[res_df$padj < 0.05 & res_df$log2FoldChange < -1] <- "Down"

# Volcano plot generated with top 10 significant genes labeled

ggplot(res_df, aes(x = log2FoldChange, y = -log10(padj), color = regulation)) +
  geom_point(alpha = 0.6) +
  scale_color_manual(
    values = c("Up" = "red", "Down" = "blue", "Not Sig" = "grey"),
    labels = c("Up" = "Overexpressed", "Down" = "Underexpressed", "Not Sig" = "Not Significant")
  ) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black") +  # Significance threshold
  geom_vline(xintercept = 0, linetype = "dotted", color = "black") +             # No fold change line
  labs(
    title = "Volcano Plot: Acidovorax vs No Bacteria (Day 14, With Aphids)",
    x = "log2 Fold Change",
    y = "-log10 adjusted p-value"
  ) +
  guides(color = guide_legend(title = "Expression")) +  # Change legend title to "Expression"
  theme_minimal()


```

Table of most sig genes: A.radicis v no bacteria day 14 aphids

```{r}
# Data ordered by adjusted p-value (padj) and top 15 most significant genes selected

top_15_genes <- res_df[order(res_df$padj), ][1:15, ]

# Gene column added from rownames
top_15_genes$Gene <- rownames(top_15_genes)

# Row names reset to prevent them from being displayed as a separate column

rownames(top_15_genes) <- NULL

# Columns reordered: Gene, log2FoldChange, padj

top_15_genes_table <- top_15_genes[, c("Gene", "log2FoldChange", "padj")]

# Values rounded for better presentation

top_15_genes_table$log2FoldChange <- round(top_15_genes_table$log2FoldChange, 2)
top_15_genes_table$padj <- round(top_15_genes_table$padj, 4)


# Table rendered as HTML

top_15_genes_table %>%
  kable("html", col.names = c("Gene", "log2 Fold Change", "Adjusted p-value")) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "responsive"),
    full_width = FALSE,
    position = "center"
  ) %>%
  column_spec(1, bold = TRUE) %>%
  column_spec(2, color = "blue") %>%
  column_spec(3, color = "red") %>%
  add_header_above(c("Top 15 Differentially Expressed Genes" = 3))  # Correct number of columns


# Ensure all of the categories appear even if one has count 0

reg_levels <- c("Up", "Down")
reg_counts <- as.data.frame(table(factor(res_df$regulation, levels = reg_levels)))
colnames(reg_counts) <- c("Regulation", "Count")

# Plotting of bar chart
ggplot(reg_counts, aes(x = Regulation, y = Count, fill = Regulation)) +
  geom_bar(stat = "identity", width = 0.6) +
  scale_fill_manual(values = c("Up" = "red", "Down" = "blue")) +
  labs(
    title = "Differentially Expressed Genes",
    x = "Regulation",
    y = "Number of Genes"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    text = element_text(size = 12)
  )

# Calculation of stats
up_count <- sum(res_df$regulation == "Up")
down_count <- sum(res_df$regulation == "Down")
total_genes <- nrow(res_df)
dysregulated_count <- up_count + down_count
dysregulated_percent <- round((dysregulated_count / total_genes) * 100, 2)

# Creation of summary table
reg_summary <- data.frame(
  Metric = c(
    "Upregulated Genes",
    "Downregulated Genes",
    "Total Genes",
    "Percent Dysregulated"
  ),
  Value = c(
    up_count,
    down_count,
    total_genes,
    paste0(dysregulated_percent, "%")
  )
)

reg_summary %>%
  kable("html", col.names = c("Metric", "Value")) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = FALSE,
    position = "center"
  ) %>%
  add_header_above(c("Differential Expression Summary" = 2))

#creation of kable extra table for clear presentation of DE summary
```

Table of most sig genes: A.radicis v b.subtilis day 1 no aphids

```{r}
# DE results for pairwise compairson

res_day7_bsub_vs_acido <- results(dds, contrast = c("condition", "Bsubtilis_day1_aphids", "Acidovorax_day1_aphids"))

# Removal of NA padj values
res_filtered <- res_day7_bsub_vs_acido[!is.na(res_day7_bsub_vs_acido$padj), ]

# Converted to dataframe & add gene names added
res_df <- as.data.frame(res_filtered)

#Gene names added to dataframe

gene_names <- rownames(counts(dds, normalized = FALSE))
res_df$gene <- gene_names[match(rownames(res_df), gene_names)]

# Regulation category added for up or downregulated genes based on log fold change

res_df$regulation <- "Not Sig"
res_df$regulation[res_df$padj < 0.05 & res_df$log2FoldChange > 1] <- "Up"
res_df$regulation[res_df$padj < 0.05 & res_df$log2FoldChange < -1] <- "Down"

# Data ordered by adjusted p-value (padj) and top 15 most significant genes selected

top15_genes <- res_df[order(res_df$padj), ][1:15, ]

# View the table
print(top15_genes)

# final table formatted
top15_table <- top15_genes[, c("gene", "log2FoldChange", "padj", "regulation")]
print(top15_table)



```

## **Volcano plot comparison: APHID V NO APHID acidovorax**

7.  ACIDOVORAX (NO APHIDS) V ACIDOVORAX (APHIDS) DAY 1

```{r}
# DE results for pairwise compairson

res_day1_acido_vs_noaphid_aphids <- results(dds, contrast = c("condition", "Acidovorax_day1_aphids", "Acidovorax_day1_no_aphids"))

# Removal of NA padj values

res_filtered <- res_day1_acido_vs_noaphid_aphids[!is.na(res_day1_acido_vs_noaphid_aphids$padj), ]

# Converted to dataframe & add gene names added

res_df <- as.data.frame(res_filtered)
res_df$gene <- rownames(res_df)

# Regulation category added for up or downregulated genes based on log fold change

res_df$regulation <- "Not Sig"
res_df$regulation[res_df$padj < 0.05 & res_df$log2FoldChange > 1] <- "Up"
res_df$regulation[res_df$padj < 0.05 & res_df$log2FoldChange < -1] <- "Down"

# Volcano plot generated with top 10 significant genes labeled

ggplot(res_df, aes(x = log2FoldChange, y = -log10(padj), color = regulation)) +
  geom_point(alpha = 0.6) +
  scale_color_manual(
    values = c("Up" = "red", "Down" = "blue", "Not Sig" = "grey"),
    labels = c("Up" = "Overexpressed", "Down" = "Underexpressed", "Not Sig" = "Not Significant")
  ) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black") +  # Significance threshold line added
  geom_vline(xintercept = 0, linetype = "dotted", color = "black") +             # shows up or down regulated 
  labs(
    title = "Volcano Plot: Acidovorax Day 1 (Aphids) vs Acidovorax Day 1 (No Aphids)",
    x = "log2 Fold Change",
    y = "-log10 adjusted p-value"
  ) +
  guides(color = guide_legend(title = "Expression")) +  # Change legend title to "Expression"
  theme_minimal()

```

8.  ACIDOVORAX (NO APHIDS) V ACIDOVORAX (APHIDS) DAY 14

```{r}
# DE results for pairwise compairson

res_day1_acido_vs_noaphid_aphids <- results(dds, contrast = c("condition", "Acidovorax_day14_aphids", "Acidovorax_day14_no_aphids"))

# Removal of NA padj values
res_filtered <- res_day1_acido_vs_noaphid_aphids[!is.na(res_day1_acido_vs_noaphid_aphids$padj), ]

# Converted to dataframe & add gene names added

res_df <- as.data.frame(res_filtered)
res_df$gene <- rownames(res_df)

# Regulation category added for up or downregulated genes based on log fold change

res_df$regulation <- "Not Sig"
res_df$regulation[res_df$padj < 0.05 & res_df$log2FoldChange > 1] <- "Up"
res_df$regulation[res_df$padj < 0.05 & res_df$log2FoldChange < -1] <- "Down"

# Volcano plot generated with top 10 significant genes labeled

ggplot(res_df, aes(x = log2FoldChange, y = -log10(padj), color = regulation)) +
  geom_point(alpha = 0.6) +
  scale_color_manual(
    values = c("Up" = "red", "Down" = "blue", "Not Sig" = "grey"),
    labels = c("Up" = "Overexpressed", "Down" = "Underexpressed", "Not Sig" = "Not Significant")
  ) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black") +  # Significance threshold line added
  geom_vline(xintercept = 0, linetype = "dotted", color = "black") +             # shows up or down regulated
  labs(
    title = "Volcano Plot: Acidovorax Day 1 (Aphids) vs Acidovorax Day 1 (No Aphids)",
    x = "log2 Fold Change",
    y = "-log10 adjusted p-value"
  ) +
  guides(color = guide_legend(title = "Expression")) +  # Change legend title to "Expression"
  theme_minimal()

```

# **Heatmap comparisons**

1.  Acidovorax v no bacteria day 1 (aphids)

```{r}
library(DESeq2)
library(pheatmap)
library(dplyr)

# Define selected samples for Acidovorax vs No Bacteria (Day 1, Aphids)
selected_samples <- c(
  "RNAseq_data_merged_1.RNA_CGR_01", "RNAseq_data_merged_2.RNA_CGR_02", 
  "RNAseq_data_merged_3.RNA_CGR_03", "RNAseq_data_merged_4.RNA_CGR_04", 
  "RNAseq_data_merged_9.RNA_CGR_9", "RNAseq_data_merged_10.RNA_CGR_10", 
  "RNAseq_data_merged_11.RNA_CGR_11", "RNAseq_data_merged_12.RNA_CGR_12"
)

# Get normalized counts from DESeqDataSet
normalized_counts <- counts(dds, normalized = TRUE)

# Subset to selected samples
normalized_counts_subset <- normalized_counts[, selected_samples]

# Remove rows with missing values
normalized_counts_subset <- normalized_counts_subset[complete.cases(normalized_counts_subset), ]

# Get top 39 DEGs based on lowest padj
top_39_genes <- res_day1_df %>%
  arrange(padj) %>%
  slice_head(n = 39) %>%
  pull(gene)

# Subset normalized counts to top 39 genes
normalized_counts_subset_top39 <- normalized_counts_subset[top_39_genes, ]

# Create annotation for sample conditions
annotation_col <- data.frame(
  condition = rep(c("Acidovorax_day1", "NoBacteria_day1"), each = 4)
)
rownames(annotation_col) <- colnames(normalized_counts_subset_top39)

# Generate heatmap
pheatmap(normalized_counts_subset_top39,
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         scale = "row",
         show_rownames = TRUE,
         show_colnames = TRUE,
         annotation_col = annotation_col,
         main = "Top 39 DEGs: Acidovorax Day 1 vs No Bacteria (Aphids)",
         fontsize = 6,
         fontsize_row = 6,
         fontsize_col = 6,
         angle_col = 45,
         filename = "heatmap_top39_deg_day1.png",
         treeheight_row = 50,
         treeheight_col = 50,
         legend = TRUE,
         mar = c(10, 10, 10, 25)
)


ncol(normalized_counts_subset_top50)
nrow(annotation_col_conditions_df)
length(annotation_col_conditions)
length(colnames(normalized_counts_subset_top50))

```

2.  Acidovorax v no bacteria day 7 (aphids)

```{r}
# Selected samples defined for Acidovorax vs No Bacteria (Day 7, Aphids)
selected_samples <- c(
  "RNAseq_data_merged_13.RNA_CGR_13", "RNAseq_data_merged_14.RNA_CGR_14", 
  "RNAseq_data_merged_15.RNA_CGR_15", "RNAseq_data_merged_16.RNA_CGR_16", 
  "RNAseq_data_merged_20.RNA_CGR_20", "RNAseq_data_merged_21.RNA_CGR_21"
)

# The counts matrix was subsetted based on selected samples
counts_subset <- counts(dds)[, selected_samples, drop =FALSE]

# Checked alignment of column and row names with selected samples
colnames(counts_subset)
rownames(counts_subset)

# Counts normalised
normalized_counts <- counts(dds, normalized = TRUE)

# Normalised counts subsetted for the selected samples
normalized_counts_subset <- normalized_counts[, selected_samples]

# Removal of rows or columns with any NA, NaN, or Inf values
normalized_counts_subset <- normalized_counts_subset[complete.cases(normalized_counts_subset), ]

# Top 50 most variable genes selected based on variance 
gene_variance <- apply(normalized_counts_subset, 1, var)
top_genes <- order(gene_variance, decreasing = TRUE)[1:50]

# Counts subsetted to only include the top 50 most variable genes
normalized_counts_subset_top50 <- normalized_counts_subset[top_genes, ]

# Condition vector manually defined to match selected samples (4 Acidovorax, 2 No Bacteria replicates)
annotation_col_conditions <- c(
  "Acidovorax_day7", "Acidovorax_day7",
  "Acidovorax_day7", "Acidovorax_day7",
  "NoBacteria_day7", "NoBacteria_day7"
)

# Condition vector converted into a dataframe
annotation_col_conditions_df <- data.frame(condition = annotation_col_conditions)

# Row names assigned to annotation data frame using column names of the heatmap data
rownames(annotation_col_conditions_df) <- colnames(normalized_counts_subset_top50)

# Structure of annotation data frame checked
str(annotation_col_conditions_df)

# Gene names added to top variable gene matrix
gene_names <- rownames(counts)
rownames(normalized_counts_subset_top50) <- gene_names[top_genes]

# Heatmap generated for Acidovorax vs No Bacteria (Day 7, Aphids)
pheatmap(normalized_counts_subset_top50,
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         scale = "row",
         show_rownames = TRUE,  # Ensured gene names shown as rownames
         show_colnames = TRUE,
         annotation_col = annotation_col_conditions_df,
         main = "Heatmap: Acidovorax Day 7 vs No Bacteria (Aphids)",
         fontsize = 6,
         fontsize_row = 6,
         fontsize_col = 6,
         angle_col = 45,  # Rotation of x-axis labels
         filename = "heatmap_acido_day7_vs_nobac_aphids.png",
         treeheight_row = 50, 
         treeheight_col = 50,
         legend = TRUE,
         mar = c(10, 10, 10, 25)  # Bottom margin increased
)


```

3.  Acidovorax v no bacteria day 14 (aphids)

```{r}
# Selected samples defined for Acidovorax vs No Bacteria (Day 14, Aphids)
selected_samples <- c(
  "RNAseq_data_merged_22.RNA_CGR_22", "RNAseq_data_merged_23.RNA_CGR_23", 
  "RNAseq_data_merged_24.RNA_CGR_24", "RNAseq_data_merged_25.RNA_CGR_25", 
  "RNAseq_data_merged_30.RNA_CGR_30", "RNAseq_data_merged_31.RNA_CGR_31", 
  "RNAseq_data_merged_32.RNA_CGR_32", "RNAseq_data_merged_33.RNA_CGR_33"
)

# The counts matrix was subsetted based on selected samples
counts_subset <- counts(dds)[, selected_samples, drop =FALSE]

# Counts normalised
normalized_counts <- counts(dds, normalized = TRUE)

# Normalised counts subsetted for the selected samples
normalized_counts_subset <- normalized_counts[, selected_samples]

# Removal of rows or columns with any NA, NaN, or Inf values
normalized_counts_subset <- normalized_counts_subset[complete.cases(normalized_counts_subset), ]

# Top 50 most variable genes selected based on variance 
gene_variance <- apply(normalized_counts_subset, 1, var)
top_genes <- order(gene_variance, decreasing = TRUE)[1:50]

# Counts subsetted to only include the top 50 most variable genes
normalized_counts_subset_top50 <- normalized_counts_subset[top_genes, ]

# Condition vector manually defined to match selected samples (4 Acidovorax, 4 No Bacteria replicates)
annotation_col_conditions <- rep(c("Acidovorax_day14", "NoBacteria_day14"), each = 4)

# Condition vector converted into a dataframe
annotation_col_conditions_df <- data.frame(condition = annotation_col_conditions)

# Structure of annotation data frame checked
str(annotation_col_conditions_df)

# Gene names added to top variable gene matrix
gene_names <- rownames(counts)
rownames(normalized_counts_subset_top50) <- gene_names[top_genes]

# Row names assigned to annotation data frame using column names of the heatmap data
rownames(annotation_col_conditions_df) <- colnames(normalized_counts_subset_top50)

# Heatmap generated for Acidovorax vs No Bacteria (Day 14, Aphids)
pheatmap(normalized_counts_subset_top50,
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         scale = "row",
         show_rownames = TRUE,  # Ensured gene names shown as rownames
         show_colnames = TRUE,
         annotation_col = annotation_col_conditions_df,
         main = "Heatmap: Acidovorax Day 14 vs No Bacteria (Aphids)",
         fontsize = 6,
         fontsize_row = 6,
         fontsize_col = 6,
         angle_col = 45,  # Rotation of x-axis labels
         filename = "heatmap_acido_day14_vs_nobac_aphids.png",
         treeheight_row = 50, 
         treeheight_col = 50,
         legend = TRUE,
         mar = c(10, 10, 10, 25)  # Bottom margin increased
)

```

4.  Acidovorax v no bacteria day 1 (no aphids)

```{r}

# Selected samples defined for Acidovorax vs No Bacteria (Day 1, No Aphids)
selected_samples <- c(
  "RNAseq_data_merged_34.RNA_CGR_34", "RNAseq_data_merged_35.RNA_CGR_35", 
  "RNAseq_data_merged_36.RNA_CGR_36", "RNAseq_data_merged_37.RNA_CGR_37", 
  "RNAseq_data_merged_42.RNA_CGR_42", "RNAseq_data_merged_43.RNA_CGR_43"
)

# The counts matrix was subsetted based on selected samples
counts_subset <- counts(dds)[, selected_samples, drop = FALSE]

# Counts normalised
normalized_counts <- counts(dds, normalized = TRUE)

# Normalised counts subsetted for the selected samples
normalized_counts_subset <- normalized_counts[, selected_samples]

# Removal of rows or columns with any NA, NaN, or Inf values
normalized_counts_subset <- normalized_counts_subset[complete.cases(normalized_counts_subset), ]

# Top 50 most variable genes selected based on variance 
gene_variance <- apply(normalized_counts_subset, 1, var)
top_genes <- order(gene_variance, decreasing = TRUE)[1:50]

# Counts subsetted to only include the top 50 most variable genes
normalized_counts_subset_top50 <- normalized_counts_subset[top_genes, ]

# Condition vector manually defined to match selected samples (4 Acidovorax, 2 No Bacteria replicates)
annotation_col_conditions <- c(
  "Acidovorax_day1", "Acidovorax_day1", "Acidovorax_day1", "Acidovorax_day1",
  "NoBacteria_day1", "NoBacteria_day1"
)

# Condition vector converted into a dataframe
annotation_col_conditions_df <- data.frame(condition = annotation_col_conditions)

# Row names assigned to annotation data frame using column names of the heatmap data
rownames(annotation_col_conditions_df) <- colnames(normalized_counts_subset_top50)

# Gene names added to top variable gene matrix
gene_names <- rownames(counts)
rownames(normalized_counts_subset_top50) <- gene_names[top_genes]

# Heatmap generated for Acidovorax vs No Bacteria (Day 1, No Aphids)
pheatmap(normalized_counts_subset_top50,
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         scale = "row",
         show_rownames = TRUE,  # Ensured gene names shown as rownames
         show_colnames = TRUE,
         annotation_col = annotation_col_conditions_df,
         main = "Heatmap: Acidovorax Day 1 vs No Bacteria (No Aphids)",
         fontsize = 6,
         fontsize_row = 6,
         fontsize_col = 6,
         angle_col = 45,  # Rotation of x-axis labels
         filename = "heatmap_acido_day1_vs_nobac_noaphids.png",
         treeheight_row = 50, 
         treeheight_col = 50,
         legend = TRUE,
         mar = c(10, 10, 10, 25)  # Bottom margin increased
)


```

5.  Acidovorax v no bacteria day 14 (no aphids)

```{r}

# Selected samples defined for Acidovorax vs No Bacteria (Day 14, No Aphids)
selected_samples <- c(
  "RNAseq_data_merged_44.RNA_CGR_44", "RNAseq_data_merged_45.RNA_CGR_45", 
  "RNAseq_data_merged_46.RNA_CGR_46",
  #"RNAseq_data_merged_50.RNA_CGR_50",  # excluded
  "RNAseq_data_merged_51.RNA_CGR_51", "RNAseq_data_merged_52.RNA_CGR_52"
  #"RNAseq_data_merged_53.RNA_CGR_53"   # excluded
)

#Samples 50 and 53 identified as outliers and therefore removed

# The counts matrix was subsetted based on selected samples
counts_subset <- counts(dds)[, selected_samples, drop = FALSE]

# Counts normalised
normalized_counts <- counts(dds, normalized = TRUE)

# Normalised counts subsetted for the selected samples
normalized_counts_subset <- normalized_counts[, selected_samples]

# Removal of rows or columns with any NA, NaN, or Inf values
normalized_counts_subset <- normalized_counts_subset[complete.cases(normalized_counts_subset), ]

# Top 50 most variable genes selected based on variance 
gene_variance <- apply(normalized_counts_subset, 1, var)
top_genes <- order(gene_variance, decreasing = TRUE)[1:50]

# Counts subsetted to only include the top 50 most variable genes
normalized_counts_subset_top50 <- normalized_counts_subset[top_genes, ]

# Condition vector manually defined to match selected samples (3 Acidovorax, 2 No Bacteria replicates)
annotation_col_conditions <- c(
  "Acidovorax_day14", "Acidovorax_day14", "Acidovorax_day14", 
  "NoBacteria_day14", "NoBacteria_day14"
)

# Condition vector converted into a dataframe
annotation_col_conditions_df <- data.frame(condition = annotation_col_conditions)

# Row names assigned to annotation data frame using column names of the heatmap data
rownames(annotation_col_conditions_df) <- colnames(normalized_counts_subset_top50)

# Gene names added to top variable gene matrix
gene_names <- rownames(counts)
rownames(normalized_counts_subset_top50) <- gene_names[top_genes]

# Heatmap generated for Acidovorax vs No Bacteria (Day 14, No Aphids)
pheatmap(normalized_counts_subset_top50,
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         scale = "row",
         show_rownames = TRUE,  # Ensured gene names shown as rownames
         show_colnames = TRUE,
         annotation_col = annotation_col_conditions_df,
         main = "Heatmap: Acidovorax Day 14 vs No Bacteria (No Aphids)",
         fontsize = 6,
         fontsize_row = 6,
         fontsize_col = 6,
         angle_col = 45,  # Rotation of x-axis labels
         filename = "heatmap_acido_day14_vs_nobac_noaphids.png",
         treeheight_row = 50, 
         treeheight_col = 50,
         legend = TRUE,
         mar = c(10, 10, 10, 25)  # Bottom margin increased
)

```

# **PCA:**

# **1. ACIDOVORAX V NO BACTERIA DAY 1 (APHIDS)**

```{r}

# Defined selected samples for PCA analysis (Day 1, Aphids)
selected_samples <- c(
  "RNAseq_data_merged_1.RNA_CGR_01", "RNAseq_data_merged_2.RNA_CGR_02", 
  "RNAseq_data_merged_3.RNA_CGR_03", "RNAseq_data_merged_4.RNA_CGR_04", 
  "RNAseq_data_merged_9.RNA_CGR_9", "RNAseq_data_merged_10.RNA_CGR_10", 
  "RNAseq_data_merged_11.RNA_CGR_11", "RNAseq_data_merged_12.RNA_CGR_12"
)

# Extracted normalized counts from DESeqDataSet
normalized_counts <- counts(dds, normalized = TRUE)

# Subsetted the normalized count matrix to include only the selected samples
normalized_counts_subset_all <- normalized_counts[, selected_samples]

# Removed genes with zero expression across all selected samples
nonzero_genes <- rowSums(normalized_counts_subset_all) > 0
normalized_counts_all_genes <- normalized_counts_subset_all[nonzero_genes, ]

# Transposed the data so that samples became rows and genes became columns (required for PCA)
pca_input <- t(normalized_counts_all_genes)

# Ran Principal Component Analysis (PCA) on scaled data
pca_result <- prcomp(pca_input, scale. = TRUE)

# Calculated percent variance explained by each principal component
percent_var <- round(100 * (pca_result$sdev^2 / sum(pca_result$sdev^2)), 1)

# Defined the condition vector corresponding to the selected samples
annotation_col_conditions <- rep(c("Acidovorax_day1", "NoBacteria_day1"), each = 4)

# Created a data frame with condition annotations and assigned row names
annotation_col_conditions_df <- data.frame(condition = annotation_col_conditions)
rownames(annotation_col_conditions_df) <- selected_samples

# Formatted the PCA result into a data frame for plotting
pca_df <- as.data.frame(pca_result$x)  # Principal component scores
pca_df$condition <- annotation_col_conditions_df$condition  # Added treatment condition
pca_df$sample <- rownames(pca_df)  # Added sample identifiers

# Loaded necessary libraries for plotting
library(ggplot2)
library(ggrepel)

# Generated the PCA plot
ggplot(pca_df, aes(x = PC1, y = PC2, color = condition)) +
  geom_point(size = 3) +  # Plotted PCA points
  xlab(paste0("PC1 (", percent_var[1], "%)")) +  # X-axis label with variance %
  ylab(paste0("PC2 (", percent_var[2], "%)")) +  # Y-axis label with variance %
  ggtitle("PCA: Acidovorax vs No Bacteria, Day 1 (Aphids)") +  # Plot title
  theme_minimal() +  # Applied a minimal theme
  theme(plot.title = element_text(hjust = 0.5))  # Centered the plot title



```

# **2. ACIDOVORAX V NO BACTERIA DAY 7 (APHIDS)**

```{r}
# Defined selected samples for Day 7 Acidovorax vs No Bacteria (Aphids)
selected_samples <- c(
  "RNAseq_data_merged_13.RNA_CGR_13", "RNAseq_data_merged_14.RNA_CGR_14", 
  "RNAseq_data_merged_15.RNA_CGR_15", "RNAseq_data_merged_16.RNA_CGR_16", 
  "RNAseq_data_merged_20.RNA_CGR_20", "RNAseq_data_merged_21.RNA_CGR_21"
)

# Retrieved normalised counts from the DESeqDataSet
normalized_counts <- counts(dds, normalized = TRUE)

# Subsetted the normalised counts matrix to include only the selected samples
normalized_counts_subset_all <- normalized_counts[, selected_samples]

# Removed genes with zero expression across all selected samples
nonzero_genes <- rowSums(normalized_counts_subset_all) > 0
normalized_counts_all_genes <- normalized_counts_subset_all[nonzero_genes, ]

# Transposed the matrix so that samples were rows and genes were columns (required for PCA)
pca_input <- t(normalized_counts_all_genes)

# Performed PCA on the scaled data
pca_result <- prcomp(pca_input, scale. = TRUE)

# Calculated the percentage of variance explained by each principal component
percent_var <- round(100 * (pca_result$sdev^2 / sum(pca_result$sdev^2)), 1)

# Created a condition vector for the selected samples
annotation_col_conditions <- c(rep("Acidovorax_day7", 4), rep("NoBacteria_day7", 2))

# Constructed an annotation data frame and assigned row names
annotation_col_conditions_df <- data.frame(condition = annotation_col_conditions)
rownames(annotation_col_conditions_df) <- selected_samples

# Converted PCA results into a data frame for plotting
pca_df <- as.data.frame(pca_result$x)
pca_df$condition <- annotation_col_conditions_df$condition
pca_df$sample <- rownames(pca_df)

# Loaded required libraries and generated the PCA plot
library(ggplot2)
library(ggrepel)

# Generated the PCA plot
ggplot(pca_df, aes(x = PC1, y = PC2, color = condition)) +
  geom_point(size = 3) +  # Plotted PCA points
  xlab(paste0("PC1 (", percent_var[1], "%)")) +  # X-axis label with variance %
  ylab(paste0("PC2 (", percent_var[2], "%)")) +  # Y-axis label with variance %
  ggtitle("PCA: Acidovorax vs No Bacteria, Day 14 Aphids") +  # Plot title
  theme_minimal() +  # Applied a minimal theme
  theme(plot.title = element_text(hjust = 0.5))  # Centered the plot title


```

# **3. ACIDOVORAX V NO BACTERIA DAY 14 (APHIDS)**

```{r}

# Defined selected samples for Day 14 Acidovorax vs No Bacteria (Aphids)
selected_samples <- c(
  "RNAseq_data_merged_22.RNA_CGR_22", "RNAseq_data_merged_23.RNA_CGR_23", 
  "RNAseq_data_merged_24.RNA_CGR_24", "RNAseq_data_merged_25.RNA_CGR_25", 
  "RNAseq_data_merged_30.RNA_CGR_30", "RNAseq_data_merged_31.RNA_CGR_31", 
  "RNAseq_data_merged_32.RNA_CGR_32", "RNAseq_data_merged_33.RNA_CGR_33"
)

# Extracted normalized counts from DESeqDataSet
normalized_counts <- counts(dds, normalized = TRUE)

# Subsetted the normalized count matrix for selected samples
normalized_counts_subset_all <- normalized_counts[, selected_samples]

# Removed genes with zero expression across all samples
nonzero_genes <- rowSums(normalized_counts_subset_all) > 0
normalized_counts_all_genes <- normalized_counts_subset_all[nonzero_genes, ]

# Transposed the data so that samples became rows and genes became columns (required for PCA)
pca_input <- t(normalized_counts_all_genes)

# Ran Principal Component Analysis (PCA) on scaled data
pca_result <- prcomp(pca_input, scale. = TRUE)

# Calculated percent variance explained by each principal component
percent_var <- round(100 * (pca_result$sdev^2 / sum(pca_result$sdev^2)), 1)

# Created condition vector for Day 14 (first 4 are Acidovorax, last 4 are NoBacteria)
annotation_col_conditions <- c(rep("Acidovorax_day14", 4), rep("NoBacteria_day14", 4))

# Created a data frame with condition annotations and assigned row names
annotation_col_conditions_df <- data.frame(condition = annotation_col_conditions)
rownames(annotation_col_conditions_df) <- selected_samples

# Formatted the PCA result into a data frame for plotting
pca_df <- as.data.frame(pca_result$x)  # Principal component scores
pca_df$condition <- annotation_col_conditions_df$condition  # Added treatment condition
pca_df$sample <- rownames(pca_df)  # Added sample identifiers

# Loaded necessary libraries for plotting
library(ggplot2)
library(ggrepel)

# Generated the PCA plot
ggplot(pca_df, aes(x = PC1, y = PC2, color = condition)) +
  geom_point(size = 3) +  # Plotted PCA points
  xlab(paste0("PC1 (", percent_var[1], "%)")) +  # X-axis label with variance %
  ylab(paste0("PC2 (", percent_var[2], "%)")) +  # Y-axis label with variance %
  ggtitle("PCA: Acidovorax vs No Bacteria, Day 14 Aphids") +  # Plot title
  theme_minimal() +  # Applied a minimal theme
  theme(plot.title = element_text(hjust = 0.5))  # Centered the plot title


```

# **4. ACIDOVORAX V NO BACTERIA (NO APHIDS) DAY 1**

```{r}
# Defined selected samples for Day 1 Acidovorax vs No Bacteria (No Aphids)
selected_samples <- c(
  "RNAseq_data_merged_34.RNA_CGR_34", "RNAseq_data_merged_35.RNA_CGR_35", 
  "RNAseq_data_merged_36.RNA_CGR_36", "RNAseq_data_merged_37.RNA_CGR_37", 
  "RNAseq_data_merged_42.RNA_CGR_42", "RNAseq_data_merged_43.RNA_CGR_43"
)

# Extracted normalized counts from DESeqDataSet
normalized_counts <- counts(dds, normalized = TRUE)

# Subsetted the normalized count matrix to include only selected samples
normalized_counts_subset_all <- normalized_counts[, selected_samples]

# Removed genes with zero expression across all selected samples
nonzero_genes <- rowSums(normalized_counts_subset_all) > 0
normalized_counts_all_genes <- normalized_counts_subset_all[nonzero_genes, ]

# Transposed the data so that samples became rows and genes became columns (required for PCA)
pca_input <- t(normalized_counts_all_genes)

# Ran Principal Component Analysis (PCA) on scaled data
pca_result <- prcomp(pca_input, scale. = TRUE)

# Calculated percent variance explained by each principal component
percent_var <- round(100 * (pca_result$sdev^2 / sum(pca_result$sdev^2)), 1)

# Created condition vector (4 Acidovorax, 2 No Bacteria for Day 1, No Aphids)
annotation_col_conditions <- c(rep("Acidovorax_day1_noaphids", 4), rep("NoBacteria_day1_noaphids", 2))

# Created a data frame with condition annotations and assigned row names
annotation_col_conditions_df <- data.frame(condition = annotation_col_conditions)
rownames(annotation_col_conditions_df) <- selected_samples

# Formatted the PCA result into a data frame for plotting
pca_df <- as.data.frame(pca_result$x)  # Principal component scores
pca_df$condition <- annotation_col_conditions_df$condition  # Added treatment condition
pca_df$sample <- rownames(pca_df)  # Added sample identifiers

# Loaded necessary libraries for plotting
library(ggplot2)
library(ggrepel)

# Generated the PCA plot
ggplot(pca_df, aes(x = PC1, y = PC2, color = condition)) +
  geom_point(size = 3) +  # Plotted PCA points
  xlab(paste0("PC1 (", percent_var[1], "%)")) +  # X-axis label with variance %
  ylab(paste0("PC2 (", percent_var[2], "%)")) +  # Y-axis label with variance %
  ggtitle("PCA: Acidovorax vs No Bacteria (Day 1, No Aphids)") +  # Plot title
  theme_minimal() +  # Applied a minimal theme
  theme(plot.title = element_text(hjust = 0.5))  # Centered the plot title

```

# **5. ACIDOVORAX V NO BACTERIA (NO APHIDS) DAY 14**

```{r}

# Defined selected samples for Day 14 Acidovorax vs No Bacteria (No Aphids)
selected_samples <- c(
  "RNAseq_data_merged_44.RNA_CGR_44", "RNAseq_data_merged_45.RNA_CGR_45", 
  "RNAseq_data_merged_46.RNA_CGR_46",
  #"RNAseq_data_merged_50.RNA_CGR_50",  # exlcuded
  "RNAseq_data_merged_51.RNA_CGR_51", "RNAseq_data_merged_52.RNA_CGR_52"
  #, "RNAseq_data_merged_53.RNA_CGR_53"  # excluded
)

# Extracted normalized counts from DESeqDataSet
normalized_counts <- counts(dds, normalized = TRUE)

# Subsetted the normalized count matrix to include only selected samples
normalized_counts_subset_all <- normalized_counts[, selected_samples]

# Removed genes with zero expression across all selected samples
nonzero_genes <- rowSums(normalized_counts_subset_all) > 0
normalized_counts_all_genes <- normalized_counts_subset_all[nonzero_genes, ]

# Transposed the data so that samples became rows and genes became columns (required for PCA)
pca_input <- t(normalized_counts_all_genes)

# Ran Principal Component Analysis (PCA) on scaled data
pca_result <- prcomp(pca_input, scale. = TRUE)

# Calculated percent variance explained by each principal component
percent_var <- round(100 * (pca_result$sdev^2 / sum(pca_result$sdev^2)), 1)

# Created condition vector (3 Acidovorax, 2 No Bacteria for Day 14, No Aphids)
annotation_col_conditions <- c(rep("Acidovorax_day14_noaphids", 3), rep("NoBacteria_day14_noaphids", 2))

# Created a data frame with condition annotations and assigned row names
annotation_col_conditions_df <- data.frame(condition = annotation_col_conditions)
rownames(annotation_col_conditions_df) <- selected_samples

# Formatted the PCA result into a data frame for plotting
pca_df <- as.data.frame(pca_result$x)  # Principal component scores
pca_df$condition <- annotation_col_conditions_df$condition  # Added treatment condition
pca_df$sample <- rownames(pca_df)  # Added sample identifiers


# Generated the PCA plot
ggplot(pca_df, aes(x = PC1, y = PC2, color = condition)) +
  geom_point(size = 3) +  # Plotted PCA points
  xlab(paste0("PC1 (", percent_var[1], "%)")) +  # X-axis label with variance %
  ylab(paste0("PC2 (", percent_var[2], "%)")) +  # Y-axis label with variance %
  ggtitle("PCA: Acidovorax vs No Bacteria (Day 14, No Aphids)") +  # Plot title
  theme_minimal() +  # Applied a minimal theme
  theme(plot.title = element_text(hjust = 0.5))  # Centered the plot title


```

# **6. ACIDOVORAX (NO APHIDS) V ACIDOVORAX (APHIDS) DAY 1**

```{r}
# Defined selected samples for Acidovorax (Aphids) vs Acidovorax (No Aphids), Day 1
selected_samples <- c(
  "RNAseq_data_merged_1.RNA_CGR_01", "RNAseq_data_merged_2.RNA_CGR_02", 
  "RNAseq_data_merged_3.RNA_CGR_03", "RNAseq_data_merged_4.RNA_CGR_04", 
  "RNAseq_data_merged_34.RNA_CGR_34", "RNAseq_data_merged_35.RNA_CGR_35", 
  "RNAseq_data_merged_36.RNA_CGR_36", "RNAseq_data_merged_37.RNA_CGR_37"
)

# Extracted normalized counts
normalized_counts <- counts(dds, normalized = TRUE)

# Subsetted normalized counts for selected samples
normalized_counts_subset_all <- normalized_counts[, selected_samples]

# Removed genes with 0 counts across all selected samples
nonzero_genes <- rowSums(normalized_counts_subset_all) > 0
normalized_counts_all_genes <- normalized_counts_subset_all[nonzero_genes, ]

# Transposed count matrix for PCA (samples as rows, genes as columns)
pca_input <- t(normalized_counts_all_genes)

# Performed PCA on scaled data
pca_result <- prcomp(pca_input, scale. = TRUE)

# Calculated percent variance explained by each principal component
percent_var <- round(100 * (pca_result$sdev^2 / sum(pca_result$sdev^2)), 1)

# Created condition vector for Acidovorax (Aphids) vs No Aphids, Day 1
annotation_col_conditions <- c(rep("Acidovorax_Aphids_day1", 4), rep("Acidovorax_NoAphids_day1", 4))

# Created annotation data frame with sample conditions
annotation_col_conditions_df <- data.frame(condition = annotation_col_conditions)
rownames(annotation_col_conditions_df) <- selected_samples

# Prepared PCA output for plotting
pca_df <- as.data.frame(pca_result$x)
pca_df$condition <- annotation_col_conditions_df$condition
pca_df$sample <- rownames(pca_df)

# Generated PCA plot
ggplot(pca_df, aes(x = PC1, y = PC2, color = condition)) +
  geom_point(size = 3) +
  xlab(paste0("PC1 (", percent_var[1], "%)")) +
  ylab(paste0("PC2 (", percent_var[2], "%)")) +
  ggtitle("PCA: Acidovorax (Aphids) vs Acidovorax (No Aphids), Day 1") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

```

# **7. ACIDOVORAX (NO APHIDS) V ACIDOVORAX (APHIDS) DAY 14**

```{r}
# Defined selected samples for Acidovorax (Aphids) vs Acidovorax (No Aphids), Day 14
selected_samples <- c(
  "RNAseq_data_merged_22.RNA_CGR_22", "RNAseq_data_merged_23.RNA_CGR_23", 
  "RNAseq_data_merged_24.RNA_CGR_24", "RNAseq_data_merged_25.RNA_CGR_25", 
  "RNAseq_data_merged_44.RNA_CGR_44", "RNAseq_data_merged_45.RNA_CGR_45", 
  "RNAseq_data_merged_46.RNA_CGR_46"
)

# Extracted normalized counts from DESeqDataSet
normalized_counts <- counts(dds, normalized = TRUE)

# Subsetted the normalized count matrix to include only selected samples
normalized_counts_subset_all <- normalized_counts[, selected_samples]

# Removed genes with zero expression across all selected samples
nonzero_genes <- rowSums(normalized_counts_subset_all) > 0
normalized_counts_all_genes <- normalized_counts_subset_all[nonzero_genes, ]

# Transposed the data so that samples became rows and genes became columns (required for PCA)
pca_input <- t(normalized_counts_all_genes)

# Performed Principal Component Analysis (PCA) on scaled data
pca_result <- prcomp(pca_input, scale. = TRUE)

# Calculated percent variance explained by each principal component
percent_var <- round(100 * (pca_result$sdev^2 / sum(pca_result$sdev^2)), 1)

# Created condition vector (4 Acidovorax + Aphids, 3 Acidovorax - Aphids) for Day 14
annotation_col_conditions <- c(rep("Acidovorax_Aphids_day14", 4), rep("Acidovorax_NoAphids_day14", 3))

# Created a data frame with condition annotations and assigned row names
annotation_col_conditions_df <- data.frame(condition = annotation_col_conditions)
rownames(annotation_col_conditions_df) <- selected_samples

# Formatted the PCA result into a data frame for plotting
pca_df <- as.data.frame(pca_result$x)  # Principal component scores
pca_df$condition <- annotation_col_conditions_df$condition  # Added treatment condition
pca_df$sample <- rownames(pca_df)  # Added sample identifiers

# Generated the PCA plot
ggplot(pca_df, aes(x = PC1, y = PC2, color = condition)) +
  geom_point(size = 3) +  # Plotted PCA points
  xlab(paste0("PC1 (", percent_var[1], "%)")) +  # X-axis label with variance %
  ylab(paste0("PC2 (", percent_var[2], "%)")) +  # Y-axis label with variance %
  ggtitle("PCA: Acidovorax (Aphids) vs Acidovorax (No Aphid), Day 14") +  # Plot title
  theme_minimal() +  # Applied a minimal theme
  theme(plot.title = element_text(hjust = 0.5))  # Centered the plot title

```

# 8. PCA: acidovorax v no bacteria day 1 (aphids vs no aphids)

```{r}
# Defined all selected samples across aphid and no-aphid treatments
selected_samples <- c(
  "RNAseq_data_merged_1.RNA_CGR_01", "RNAseq_data_merged_2.RNA_CGR_02", 
  "RNAseq_data_merged_3.RNA_CGR_03", "RNAseq_data_merged_4.RNA_CGR_04", 
  "RNAseq_data_merged_9.RNA_CGR_9", "RNAseq_data_merged_10.RNA_CGR_10", 
  "RNAseq_data_merged_11.RNA_CGR_11", "RNAseq_data_merged_12.RNA_CGR_12",
  "RNAseq_data_merged_34.RNA_CGR_34", "RNAseq_data_merged_35.RNA_CGR_35", 
  "RNAseq_data_merged_36.RNA_CGR_36", "RNAseq_data_merged_37.RNA_CGR_37", 
  "RNAseq_data_merged_42.RNA_CGR_42", "RNAseq_data_merged_43.RNA_CGR_43"
)

# Extracted normalized counts from DESeqDataSet
normalized_counts <- counts(dds, normalized = TRUE)

# Subsetted the normalized count matrix to include only selected samples
normalized_counts_subset_all <- normalized_counts[, selected_samples]

# Filtered out low-expressed genes: kept genes with at least 10 counts in at least 3 samples
keep_genes <- rowSums(normalized_counts_subset_all >= 10) >= 3
normalized_counts_filtered <- normalized_counts_subset_all[keep_genes, ]

# Selected top 500 most variable genes across samples for PCA
top_var_genes <- head(order(apply(normalized_counts_filtered, 1, var), decreasing = TRUE), 500)

# Transposed the matrix to have samples as rows and selected genes as columns
pca_input <- t(normalized_counts_filtered[top_var_genes, ])

# Performed Principal Component Analysis (PCA) on scaled expression data
pca_result <- prcomp(pca_input, scale. = TRUE)

# Calculated percent variance explained by each principal component
percent_var <- round(100 * (pca_result$sdev^2 / sum(pca_result$sdev^2)), 1)

# Defined treatment group for each sample (Acidovorax vs No Bacteria)
treatment <- c(rep("Acidovorax", 4), rep("NoBacteria", 4), rep("Acidovorax", 4), rep("NoBacteria", 2))

# Defined aphid presence condition for each sample
aphids <- c(rep("Aphids", 8), rep("No Aphids", 6))

# Created a data frame with PCA scores and sample metadata
pca_df <- as.data.frame(pca_result$x)  # Principal component scores
pca_df$treatment <- treatment  # Added treatment information
pca_df$aphids <- aphids  # Added aphid presence condition
pca_df$sample <- selected_samples  # Added sample IDs

# Loaded necessary libraries for plotting
library(ggplot2)
library(ggrepel)

# Plotted PCA: colored by aphid presence, shaped by treatment type
ggplot(pca_df, aes(x = PC1, y = PC2, color = aphids, shape = treatment)) +
  geom_point(size = 3) +  # Plot points with size
  scale_shape_manual(values = c("Acidovorax" = 16, "NoBacteria" = 17)) +  # Circle for Acidovorax, triangle for NoBacteria
  xlab(paste0("PC1 (", percent_var[1], "%)")) +  # X-axis with PC1 variance
  ylab(paste0("PC2 (", percent_var[2], "%)")) +  # Y-axis with PC2 variance
  ggtitle("PCA: Day 1  Aphid Presence and Treatment") +  # Plot title
  theme_minimal() +  # Clean theme
  theme(
    plot.title = element_text(hjust = 0.5),  # Centered title
    legend.title = element_blank()  # Removed legend title
  )

```

# 9. PCA: acidovorax v no bacteria day 14 (aphids vs no aphids)

```{r}
# Combined Day 14 sample names including both Aphids and No Aphids groups
selected_samples <- c(
  "RNAseq_data_merged_22.RNA_CGR_22", "RNAseq_data_merged_23.RNA_CGR_23", 
  "RNAseq_data_merged_24.RNA_CGR_24", "RNAseq_data_merged_25.RNA_CGR_25", 
  "RNAseq_data_merged_30.RNA_CGR_30", "RNAseq_data_merged_31.RNA_CGR_31", 
  "RNAseq_data_merged_32.RNA_CGR_32", "RNAseq_data_merged_33.RNA_CGR_33",
  "RNAseq_data_merged_44.RNA_CGR_44", "RNAseq_data_merged_45.RNA_CGR_45", 
  "RNAseq_data_merged_46.RNA_CGR_46",
  #"RNAseq_data_merged_50.RNA_CGR_50",  # Excluded sample
  "RNAseq_data_merged_51.RNA_CGR_51", "RNAseq_data_merged_52.RNA_CGR_52"
  #, "RNAseq_data_merged_53.RNA_CGR_53"  # Optional sample
)

# Extracted normalized counts from DESeqDataSet object
normalized_counts <- counts(dds, normalized = TRUE)

# Subsetted the normalized counts matrix to retain only selected samples
normalized_counts_subset_all <- normalized_counts[, selected_samples]

# Filtered out low-expressed genes: retained those with at least 10 counts in at least 3 samples
keep_genes <- rowSums(normalized_counts_subset_all >= 10) >= 3
normalized_counts_filtered <- normalized_counts_subset_all[keep_genes, ]

# Selected the top 500 most variable genes for PCA analysis
top_var_genes <- head(order(apply(normalized_counts_filtered, 1, var), decreasing = TRUE), 500)

# Transposed matrix: samples as rows, genes as columns
pca_input <- t(normalized_counts_filtered[top_var_genes, ])

# Performed PCA on scaled gene expression data
pca_result <- prcomp(pca_input, scale. = TRUE)

# Calculated the percentage of variance explained by the first few principal components
percent_var <- round(100 * (pca_result$sdev^2 / sum(pca_result$sdev^2)), 1)

# Defined sample treatment conditions (Acidovorax vs No Bacteria)
treatment <- c(rep("Acidovorax", 4), rep("NoBacteria", 4), rep("Acidovorax", 3), rep("NoBacteria", 2))

# Defined aphid presence condition (Aphids vs No Aphids)
aphids <- c(rep("Aphids", 8), rep("No Aphids", 5))

# Created data frame with PCA scores and metadata
pca_df <- as.data.frame(pca_result$x)  # PCA scores
pca_df$treatment <- treatment  # Treatment condition per sample
pca_df$aphids <- aphids  # Aphid presence per sample
pca_df$sample <- selected_samples  # Sample IDs

# Loaded required libraries for plotting (ensure they are already loaded in your session)
library(ggplot2)
library(ggrepel)

# Plotted PCA: colored by aphid presence, shaped by treatment type
ggplot(pca_df, aes(x = PC1, y = PC2, color = aphids, shape = treatment)) +
  geom_point(size = 3) +  # Plot points with size
  scale_shape_manual(values = c("Acidovorax" = 16, "NoBacteria" = 17)) +  # Circle for Acidovorax, triangle for NoBacteria
  xlab(paste0("PC1 (", percent_var[1], "%)")) +  # X-axis with PC1 variance
  ylab(paste0("PC2 (", percent_var[2], "%)")) +  # Y-axis with PC2 variance
  ggtitle("PCA: Day 14  Aphid Presence and Treatment") +  # Plot title
  theme_minimal() +  # Clean theme
  theme(
    plot.title = element_text(hjust = 0.5),  # Centered title
    legend.title = element_blank()  # Removed legend title
  )




```

# 10. PCA: ALL SAMPLES by aphid presence

```{r}
# Loaded required library
library(ggplot2)

# Defined selected samples across all conditions and time points
selected_samples <- c(
  # Acidovorax + Aphids
  "RNAseq_data_merged_1.RNA_CGR_01", "RNAseq_data_merged_2.RNA_CGR_02",
  "RNAseq_data_merged_3.RNA_CGR_03", "RNAseq_data_merged_4.RNA_CGR_04",       
  "RNAseq_data_merged_13.RNA_CGR_13", "RNAseq_data_merged_14.RNA_CGR_14",
  "RNAseq_data_merged_15.RNA_CGR_15", "RNAseq_data_merged_16.RNA_CGR_16",   
  "RNAseq_data_merged_22.RNA_CGR_22", "RNAseq_data_merged_23.RNA_CGR_23",
  "RNAseq_data_merged_24.RNA_CGR_24", "RNAseq_data_merged_25.RNA_CGR_25",   

  # Acidovorax + No Aphids
  "RNAseq_data_merged_34.RNA_CGR_34", "RNAseq_data_merged_35.RNA_CGR_35",
  "RNAseq_data_merged_36.RNA_CGR_36", "RNAseq_data_merged_37.RNA_CGR_37",   
  "RNAseq_data_merged_44.RNA_CGR_44", "RNAseq_data_merged_45.RNA_CGR_45",
  "RNAseq_data_merged_46.RNA_CGR_46",                                       

  # No Bacteria + Aphids
  "RNAseq_data_merged_9.RNA_CGR_9", "RNAseq_data_merged_10.RNA_CGR_10",
  "RNAseq_data_merged_11.RNA_CGR_11", "RNAseq_data_merged_12.RNA_CGR_12",   
  "RNAseq_data_merged_20.RNA_CGR_20", "RNAseq_data_merged_21.RNA_CGR_21",   
  "RNAseq_data_merged_30.RNA_CGR_30", "RNAseq_data_merged_31.RNA_CGR_31",
  "RNAseq_data_merged_32.RNA_CGR_32", "RNAseq_data_merged_33.RNA_CGR_33",   

  # No Bacteria + No Aphids
  "RNAseq_data_merged_42.RNA_CGR_42", "RNAseq_data_merged_43.RNA_CGR_43", "RNAseq_data_merged_51.RNA_CGR_51",
  "RNAseq_data_merged_52.RNA_CGR_52"    
)

# Defined sample metadata
treatment <- c(rep("Acidovorax", 19), rep("NoBacteria", 14))
aphids <- c(rep("Aphids", 12), rep("NoAphids", 7), rep("Aphids", 10), rep("NoAphids", 4))
day <- c(
  rep("Day 1", 4), rep("Day 7", 4), rep("Day 14", 4),  # Acidovorax + Aphids
  rep("Day 1", 4), rep("Day 14", 3),                   # Acidovorax + No Aphids
  rep("Day 1", 4), rep("Day 7", 2), rep("Day 14", 4),  # No Bacteria + Aphids
  rep("Day 1", 2), rep("Day 14", 2)                    # No Bacteria + No Aphids
)

# Extracted normalized counts from DESeq2 object
normalized_counts <- counts(dds, normalized = TRUE)

# Subsetted the counts matrix to selected samples
normalized_counts_subset <- normalized_counts[, selected_samples]

# Removed genes with zero counts across all samples
nonzero_genes <- rowSums(normalized_counts_subset) > 0
normalized_counts_filtered <- normalized_counts_subset[nonzero_genes, ]

# Transposed the matrix and performed PCA
pca_input <- t(normalized_counts_filtered)
pca_result <- prcomp(pca_input, scale. = TRUE)

# Calculated percent variance explained by PC1 and PC2
percent_var <- round(100 * (pca_result$sdev^2 / sum(pca_result$sdev^2)), 1)

# Created a data frame containing PCA coordinates and metadata
pca_df <- as.data.frame(pca_result$x)
pca_df$treatment <- treatment
pca_df$aphids <- aphids
pca_df$day <- factor(day, levels = c("Day 1", "Day 7", "Day 14"))
pca_df$sample <- rownames(pca_df)

# Generated final PCA plot: colored by aphid status, shaped by treatment
ggplot(pca_df, aes(x = PC1, y = PC2, color = aphids, shape = treatment, label = sample)) +
  geom_point(size = 3) +
  scale_color_manual(values = c("Aphids" = "purple", "NoAphids" = "orange")) +
  scale_shape_manual(values = c("Acidovorax" = 16, "NoBacteria" = 15)) +  # Circle = Acidovorax, Square = NoBacteria
  xlab(paste0("PC1 (", percent_var[1], "%)")) +
  ylab(paste0("PC2 (", percent_var[2], "%)")) +
  ggtitle("PCA: Aphid Status and Treatment") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))



```

# 11. PCA all samples: By condition

```{r}
# Loaded required library
library(ggplot2)

# Defined selected samples across treatment and aphid conditions
selected_samples <- c(
  # Acidovorax + Aphids
  "RNAseq_data_merged_1.RNA_CGR_01", "RNAseq_data_merged_2.RNA_CGR_02",
  "RNAseq_data_merged_3.RNA_CGR_03", "RNAseq_data_merged_4.RNA_CGR_04",       
  "RNAseq_data_merged_13.RNA_CGR_13", "RNAseq_data_merged_14.RNA_CGR_14",
  "RNAseq_data_merged_15.RNA_CGR_15", "RNAseq_data_merged_16.RNA_CGR_16",   
  "RNAseq_data_merged_22.RNA_CGR_22", "RNAseq_data_merged_23.RNA_CGR_23",
  "RNAseq_data_merged_24.RNA_CGR_24", "RNAseq_data_merged_25.RNA_CGR_25",   

  # Acidovorax + No Aphids
  "RNAseq_data_merged_34.RNA_CGR_34", "RNAseq_data_merged_35.RNA_CGR_35",
  "RNAseq_data_merged_36.RNA_CGR_36", "RNAseq_data_merged_37.RNA_CGR_37",   
  "RNAseq_data_merged_44.RNA_CGR_44", "RNAseq_data_merged_45.RNA_CGR_45",
  "RNAseq_data_merged_46.RNA_CGR_46",                                       

  # No Bacteria + Aphids
  "RNAseq_data_merged_9.RNA_CGR_9", "RNAseq_data_merged_10.RNA_CGR_10",
  "RNAseq_data_merged_11.RNA_CGR_11", "RNAseq_data_merged_12.RNA_CGR_12",   
  "RNAseq_data_merged_20.RNA_CGR_20", "RNAseq_data_merged_21.RNA_CGR_21",   
  "RNAseq_data_merged_30.RNA_CGR_30", "RNAseq_data_merged_31.RNA_CGR_31",
  "RNAseq_data_merged_32.RNA_CGR_32", "RNAseq_data_merged_33.RNA_CGR_33",   

  # No Bacteria + No Aphids
  "RNAseq_data_merged_42.RNA_CGR_42", "RNAseq_data_merged_43.RNA_CGR_43", 
  "RNAseq_data_merged_51.RNA_CGR_51", "RNAseq_data_merged_52.RNA_CGR_52"
)

# Defined metadata vectors
day <- c(
  rep("Day 1", 4), rep("Day 7", 4), rep("Day 14", 4),   # Acidovorax + Aphids
  rep("Day 1", 4), rep("Day 14", 3),                    # Acidovorax + No Aphids
  rep("Day 1", 4), rep("Day 7", 2), rep("Day 14", 4),   # No Bacteria + Aphids
  rep("Day 1", 2), rep("Day 14", 2)                     # No Bacteria + No Aphids
)

treatment <- c(
  rep("Acidovorax", 19),
  rep("NoBacteria", 14)
)

aphids <- c(
  rep("Aphids", 12),
  rep("NoAphids", 7),
  rep("Aphids", 10),
  rep("NoAphids", 4)
)

# Created a combined group identifier
group <- paste(treatment, aphids, sep = "_")

# Retrieved normalized counts from DESeq2 object and subset to selected samples
normalized_counts <- counts(dds, normalized = TRUE)
normalized_counts_subset <- normalized_counts[, selected_samples]

# Filtered out genes with zero total counts across all selected samples
nonzero_genes <- rowSums(normalized_counts_subset) > 0
normalized_counts_filtered <- normalized_counts_subset[nonzero_genes, ]

# Transposed matrix for PCA and ran PCA
pca_input <- t(normalized_counts_filtered)
pca_result <- prcomp(pca_input, scale. = TRUE)

# Calculated percent variance explained for PC1 and PC2
percent_var <- round(100 * (pca_result$sdev^2 / sum(pca_result$sdev^2)), 1)

# Constructed PCA dataframe with metadata
pca_df <- as.data.frame(pca_result$x)
pca_df$treatment <- treatment
pca_df$aphids <- aphids
pca_df$day <- factor(day, levels = c("Day 1", "Day 7", "Day 14"))
pca_df$group <- group
pca_df$sample <- rownames(pca_df)

# Generated PCA plot: color by group, shape by day
ggplot(pca_df, aes(x = PC1, y = PC2, color = group, shape = day)) +
  geom_point(size = 3) +
  scale_shape_manual(values = c(16, 17, 15)) +  # Circle = Day 1, Triangle = Day 7, Square = Day 14
  scale_color_manual(values = c(
    "Acidovorax_Aphids" = "red",
    "Acidovorax_NoAphids" = "red",
    "NoBacteria_Aphids" = "blue",
    "NoBacteria_NoAphids" = "blue"
  )) +
  xlab(paste0("PC1 (", percent_var[1], "%)")) +
  ylab(paste0("PC2 (", percent_var[2], "%)")) +
  ggtitle("PCA: Acidovorax vs No Bacteria by Aphid Status and Day") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

```

# 12. PCA all samples + B.subtillis samples

```{r}
# Defined selected samples across treatment and aphid conditions
selected_samples <- c(
  # Acidovorax + Aphids
  "RNAseq_data_merged_1.RNA_CGR_01", "RNAseq_data_merged_2.RNA_CGR_02",
  "RNAseq_data_merged_3.RNA_CGR_03", "RNAseq_data_merged_4.RNA_CGR_04",       
  "RNAseq_data_merged_13.RNA_CGR_13", "RNAseq_data_merged_14.RNA_CGR_14",
  "RNAseq_data_merged_15.RNA_CGR_15", "RNAseq_data_merged_16.RNA_CGR_16",   
  "RNAseq_data_merged_22.RNA_CGR_22", "RNAseq_data_merged_23.RNA_CGR_23",
  "RNAseq_data_merged_24.RNA_CGR_24", "RNAseq_data_merged_25.RNA_CGR_25",   

  # Acidovorax + No Aphids
  "RNAseq_data_merged_34.RNA_CGR_34", "RNAseq_data_merged_35.RNA_CGR_35",
  "RNAseq_data_merged_36.RNA_CGR_36", "RNAseq_data_merged_37.RNA_CGR_37",   
  "RNAseq_data_merged_44.RNA_CGR_44", "RNAseq_data_merged_45.RNA_CGR_45",
  "RNAseq_data_merged_46.RNA_CGR_46",                                       

  # No Bacteria + Aphids
  "RNAseq_data_merged_9.RNA_CGR_9", "RNAseq_data_merged_10.RNA_CGR_10",
  "RNAseq_data_merged_11.RNA_CGR_11", "RNAseq_data_merged_12.RNA_CGR_12",   
  "RNAseq_data_merged_20.RNA_CGR_20", "RNAseq_data_merged_21.RNA_CGR_21",   
  "RNAseq_data_merged_30.RNA_CGR_30", "RNAseq_data_merged_31.RNA_CGR_31",
  "RNAseq_data_merged_32.RNA_CGR_32", "RNAseq_data_merged_33.RNA_CGR_33",   

  # No Bacteria + No Aphids
  "RNAseq_data_merged_42.RNA_CGR_42", "RNAseq_data_merged_43.RNA_CGR_43", 
  "RNAseq_data_merged_51.RNA_CGR_51", "RNAseq_data_merged_52.RNA_CGR_52",

  # B. subtilis + Aphids (excluding 5 and 28)
  "RNAseq_data_merged_6.RNA_CGR_6", "RNAseq_data_merged_7.RNA_CGR_7", "RNAseq_data_merged_8.RNA_CGR_8",
  "RNAseq_data_merged_17.RNA_CGR_17", "RNAseq_data_merged_18.RNA_CGR_18", "RNAseq_data_merged_19.RNA_CGR_19",
  "RNAseq_data_merged_26.RNA_CGR_26", "RNAseq_data_merged_27.RNA_CGR_27", "RNAseq_data_merged_29.RNA_CGR_29",

  # B. subtilis + No Aphids
  "RNAseq_data_merged_38.RNA_CGR_38", "RNAseq_data_merged_39.RNA_CGR_39",
  "RNAseq_data_merged_40.RNA_CGR_40", "RNAseq_data_merged_41.RNA_CGR_41",
  "RNAseq_data_merged_47.RNA_CGR_47", "RNAseq_data_merged_48.RNA_CGR_48",
  "RNAseq_data_merged_49.RNA_CGR_49"
)

# Assigned sample metadata
day <- c(
  rep("Day 1", 4), rep("Day 7", 4), rep("Day 14", 4),   
  rep("Day 1", 4), rep("Day 14", 3),                    
  rep("Day 1", 4), rep("Day 7", 2), rep("Day 14", 4),   
  rep("Day 1", 2), rep("Day 14", 2),                    
  rep("Day 1", 3), rep("Day 7", 3), rep("Day 14", 3),   
  rep("Day 1", 4), rep("Day 14", 3)                     
)

treatment <- c(
  rep("Acidovorax", 19),
  rep("NoBacteria", 14),
  rep("Bsubtilis", 16)
)

aphids <- c(
  rep("Aphids", 12),           
  rep("NoAphids", 7),          
  rep("Aphids", 10),           
  rep("NoAphids", 4),          
  rep("Aphids", 9),            
  rep("NoAphids", 7)           
)

# Created combined group identifier
group <- paste(treatment, aphids, sep = "_")

# Defined color mapping for groups
color_scale <- c(
  "Acidovorax_Aphids" = "red",
  "Acidovorax_NoAphids" = "orange",
  "NoBacteria_Aphids" = "blue",
  "NoBacteria_NoAphids" = "purple",
  "Bsubtilis_Aphids" = "darkgreen",
  "Bsubtilis_NoAphids" = "lightgreen"
)

# Extracted normalized counts and subsetted the matrix
normalized_counts <- counts(dds, normalized = TRUE)
normalized_counts_subset <- normalized_counts[, selected_samples]

# Removed genes with zero total counts
nonzero_genes <- rowSums(normalized_counts_subset) > 0
normalized_counts_filtered <- normalized_counts_subset[nonzero_genes, ]

# Transposed and scaled the matrix for PCA
pca_input <- t(normalized_counts_filtered)
pca_result <- prcomp(pca_input, scale. = TRUE)
percent_var <- round(100 * (pca_result$sdev^2 / sum(pca_result$sdev^2)), 1)

# Constructed PCA data frame with metadata
pca_df <- as.data.frame(pca_result$x)
pca_df$treatment <- treatment
pca_df$aphids <- aphids
pca_df$day <- factor(day, levels = c("Day 1", "Day 7", "Day 14"))
pca_df$group <- group
pca_df$sample <- rownames(pca_df)

library(ggrepel)

ggplot(pca_df, aes(x = PC1, y = PC2, color = group, shape = day)) +
  geom_point(size = 3) +
  scale_shape_manual(values = c(16, 17, 15)) +
  scale_color_manual(values = color_scale) +
  xlab(paste0("PC1 (", percent_var[1], "%)")) +
  ylab(paste0("PC2 (", percent_var[2], "%)")) +
  ggtitle("PCA: Treatment and Aphid Status Across Time") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))


```

> ```{r}
> >
> # Define comparisons for day 14
> comparisons <- list(
>   c("Acidovorax_day14_no_aphids", "Bsubtilis_day14_no_aphids"),
>   c("Acidovorax_day14_no_aphids", "NoBacteria_day14_no_aphids"),
>   c("Bsubtilis_day14_no_aphids", "NoBacteria_day14_no_aphids")
> )
> >
> # Store DEG counts
> deg_counts <- data.frame(Treatment = character(), Direction = character(), Num_DEGs = numeric())
> >
> # Loop over each pairwise comparison
> for (cmp in comparisons) {
>   res <- results(dds, contrast = c("condition", cmp[1], cmp[2]))
>   res <- res[!is.na(res$padj), ]
> >
>   up <- res[res$padj < 1 & res$log2FoldChange > 1, ]
>   down <- res[res$padj < 1 & res$log2FoldChange < -1, ]
> >
>   # Create a label for the treatment
>   treatment_label <- paste(gsub("_day14_no_aphids", "", cmp[1]), "vs", gsub("_day14_no_aphids", "", cmp[2]))
> >
>   # Append results to the DEG counts data frame
>   deg_counts <- rbind(
>     deg_counts, 
>     data.frame(Treatment = treatment_label, Direction = "Upregulated", Num_DEGs = nrow(up)),
>     data.frame(Treatment = treatment_label, Direction = "Downregulated", Num_DEGs = nrow(down))
>   )
> }
> >
> # View results
> print(deg_counts)
> >
> ```

## **Metatranscriptomics Aphid Project (MAPPED TO B.subtilis)**

```{r}
#Original code without ordering / gene names
# Load GFF
gff_file <- '/Users/alexbeggs/Dropbox/Aphid_project_703/RNA_SEQ_.B_subtilis/252424E_Bacillussubtilis_trimmed.gff'
df <- read.gff(gff_file)

# Keep only CDS features
df <- df[df$type == "CDS", ]

# Extract attributes
ID <- rep(NA, nrow(df))
locus_tag <- rep(NA, nrow(df))
product <- rep(NA, nrow(df))
widths <- df$end - df$start + 1

for(k in 1:nrow(df)){
  tmp <- unlist(strsplit(df$attributes[k], split = ';'))
  for(i in seq_along(tmp)){
    tmp2 <- unlist(strsplit(tmp[i], split = '='))
    if(length(tmp2) == 2){
      if(tmp2[1] == 'ID') ID[k] <- tmp2[2]
      if(tmp2[1] == 'locus_tag') locus_tag[k] <- tmp2[2]
      if(tmp2[1] == 'product') product[k] <- tmp2[2]
    }
  }
}

# If seqid is just '1', leave it as-is (don't add long label unless BAMs require it)
df$seqid <- as.vector(df$seqid)

# Build SB dataframe
SB <- data.frame(
  strain = rep('Bsubtilis', nrow(df)),
  seqid = df$seqid,
  type = df$type,
  ID = ID,
  locus_tag = locus_tag,
  product = product,
  strand = df$strand,
  start = df$start,
  end = df$end,
  widths = widths
)

# Create GRanges object
featuresB <- GRanges(
  seqnames = SB$seqid,
  ranges = IRanges(start = SB$start, end = SB$end),
  strand = SB$strand,
  group_id = SB[, c(1, 3:5)]
)

# Set working directory with BAM files
setwd('/Users/alexbeggs/Dropbox/Aphid_project_703/RNA_SEQ_.B_subtilis/BAM_files_B.subtilis')

# Get list of BAM files
bamfiles <- list.files(recursive = TRUE, pattern = "*bam$", full.names = TRUE)

# Summarize overlaps (set singleEnd=FALSE if your reads are paired-end!)
olap <- summarizeOverlaps(featuresB, bamfiles, mode = 'Union')

# Extract count matrix
counts <- assay(olap)

# Initialize TPM matrix
TPMs <- matrix(NA, nrow(counts), ncol(counts))
ind <- 1:nrow(counts)

# Compute TPMs
for(k in 1:ncol(counts)){
  tmp <- counts[ind, k] / (SB$widths[ind] / 1000) 
  pmF <- sum(tmp) / 1e6
  TPMs[ind, k] <- tmp / pmF
}

# Sample names from BAMs
sample_names <- sub("\\.bam$", "", basename(bamfiles))
colnames(counts) <- sample_names
colnames(TPMs) <- sample_names

# Assign gene names or fallback to ID
gene_labels <- ifelse(is.na(SB$locus_tag), SB$ID, SB$locus_tag)
rownames(counts) <- make.unique(gene_labels)
rownames(TPMs) <- rownames(counts)

# Save output
write.csv(counts, 'raw_counts_b.subtilis.csv', row.names = TRUE)
write.csv(TPMs, 'TPMs_b.subtilis.csv', row.names = TRUE)

```

```{r}
#Sorting SB

gff_file <- '/Users/alexbeggs/Dropbox/Aphid_project_703/RNA_SEQ_.B_subtilis/252424E_Bacillussubtilis_trimmed.gff'

df <- read.gff(gff_file)
ID = rep(NA, nrow(df))
locus_tag = rep(NA, nrow(df))
product = rep(NA, nrow(df))
gene_name = rep(NA, nrow(df))  # Added for gene name
widths <- df$end - df$start + 1

# Extract relevant attributes from the GFF file
for(k in 1:nrow(df)){
  tmp <- unlist(strsplit(df$attributes[k], split = ';'))
  for (i in 1:length(tmp)){
    tmp2 = unlist(strsplit(tmp[i], split = '='))
    if(tmp2[1] == 'ID') ID[k] <- tmp2[2]
    if(tmp2[1] == 'locus_tag') locus_tag[k] <- tmp2[2]
    if(tmp2[1] == 'product') product[k] <- tmp2[2]
    if(tmp2[1] == 'gene') gene_name[k] <- tmp2[2]  # Extract gene name
  }
}

# Check if gene names are missing and replace with ID or locus_tag
gene_name[is.na(gene_name)] <- ID[is.na(gene_name)]  # Fallback to ID if gene_name is NA
# Alternatively, you can use locus_tag instead of ID:
# gene_name[is.na(gene_name)] <- locus_tag[is.na(gene_name)]

# Fix seqid
df$seqid <- as.vector(df$seqid)
df$seqid[df$seqid == '1'] <- '1 length=4286715 depth=1.00x circular=true'

# Create SB dataframe with gene names
SB = data.frame(
  strain = rep('Bsubtilis', nrow(df)),
  seqid = df$seqid,
  type = df$type,
  ID,
  locus_tag,
  product,
  gene_name,  # Add gene_name to the dataframe
  strand = df$strand,
  start = df$start,
  end = df$end,
  widths = widths
)

# Create GRanges object
featuresB <- GRanges(
  seqnames = SB$seqid,
  ranges = IRanges(start = SB$start, end = SB$end),
  strand = SB$strand,
  group_id = SB[, c(1, 3:5)]
)

setwd('/Users/alexbeggs/Dropbox/Aphid_project_703/RNA_SEQ_.B_subtilis/BAM_files_B.subtilis')

# Get list of BAM files
bamfiles <- list.files(recursive = TRUE, pattern = "*bam$", full.names = TRUE)

# Summarize overlaps
olap <- summarizeOverlaps(featuresB, bamfiles, mode = 'Union')

# Extract count matrix
counts_bsubtilis <- assay(olap)

# Initialize TPM matrix
TPMs_bsubtilis <- matrix(NA, nrow(counts_bsubtilis), ncol(counts_bsubtilis))


# Define 'ind' as all row indices
ind <- 1:nrow(counts)

# Compute TPMs
for(k in 1:ncol(counts)){
  tmp <- counts[ind, k] / (SB$widths[ind] / 1000) 
  pmF <- sum(tmp) / 1000000
  TPM <- tmp / pmF
  TPMs[ind, k] <- TPM
}

# Extract sample names from BAM filenames
sample_names <- basename(bamfiles)  # Extracts just the filename
sample_names <- sub("\\.bam$", "", sample_names)  # Removes ".bam" extension

# Assign column names to the count and TPM matrices
colnames(counts) <- sample_names
colnames(TPMs) <- sample_names

# Ensure row names (gene names) are included
rownames(counts) <- SB$gene_name  # Use gene names
rownames(TPMs) <- SB$gene_name    # Use gene names

# Write to CSV including row and column names
write.csv(counts, 'raw_counts_b.subtilis.csv', row.names = TRUE)
write.csv(TPMs, 'TPMs_b.subtilis.csv', row.names = TRUE)

# Define the desired sample order by actual sample names (based on your specification)
desired_order <- c(
  "RNAseq_data_merged_1-RNA_CGR_01", "RNAseq_data_merged_2-RNA_CGR_02", "RNAseq_data_merged_3-RNA_CGR_03", "RNAseq_data_merged_4-RNA_CGR_04",    # acidovorax day 1 aphids
  "RNAseq_data_merged_13-RNA_CGR_13", "RNAseq_data_merged_14-RNA_CGR_14", "RNAseq_data_merged_15-RNA_CGR_15", "RNAseq_data_merged_16-RNA_CGR_16",  # acidovorax day 7 aphids
  "RNAseq_data_merged_22-RNA_CGR_22", "RNAseq_data_merged_23-RNA_CGR_23", "RNAseq_data_merged_24-RNA_CGR_24", "RNAseq_data_merged_25-RNA_CGR_25",  # acidovorax day 14 aphids
  "RNAseq_data_merged_34-RNA_CGR_34", "RNAseq_data_merged_35-RNA_CGR_35", "RNAseq_data_merged_36-RNA_CGR_36", "RNAseq_data_merged_37-RNA_CGR_37",  # acidovorax day 1 no aphids
  "RNAseq_data_merged_44-RNA_CGR_44", "RNAseq_data_merged_45-RNA_CGR_45", "RNAseq_data_merged_46-RNA_CGR_46",                            # acidovorax day 14 no aphids
  "RNAseq_data_merged_9-RNA_CGR_9", "RNAseq_data_merged_10-RNA_CGR_10", "RNAseq_data_merged_11-RNA_CGR_11", "RNAseq_data_merged_12-RNA_CGR_12",  # No bacteria day 1 aphids
  "RNAseq_data_merged_20-RNA_CGR_20", "RNAseq_data_merged_21-RNA_CGR_21",                                                  # No bacteria day 7 aphids
  "RNAseq_data_merged_30-RNA_CGR_30", "RNAseq_data_merged_31-RNA_CGR_31", "RNAseq_data_merged_32-RNA_CGR_32", "RNAseq_data_merged_33-RNA_CGR_33",  # No bacteria day 14 aphids
  "RNAseq_data_merged_42-RNA_CGR_42", "RNAseq_data_merged_43-RNA_CGR_43",                                                  # No bacteria day 1 no aphids
  "RNAseq_data_merged_50-RNA_CGR_50", "RNAseq_data_merged_51-RNA_CGR_51", "RNAseq_data_merged_52-RNA_CGR_52", "RNAseq_data_merged_53-RNA_CGR_53",   # No bacteria day 14 no aphids
  "RNAseq_data_merged_5-RNA_CGR_5", "RNAseq_data_merged_6-RNA_CGR_6", "RNAseq_data_merged_7-RNA_CGR_7", "RNAseq_data_merged_8-RNA_CGR_8",    # B.Subtilis day 1 aphids
  "RNAseq_data_merged_17-RNA_CGR_17", "RNAseq_data_merged_18-RNA_CGR_18", "RNAseq_data_merged_19-RNA_CGR_19",                            # B.Subtilis day 7 aphids
  "RNAseq_data_merged_26-RNA_CGR_26", "RNAseq_data_merged_27-RNA_CGR_27", "RNAseq_data_merged_28-RNA_CGR_28", "RNAseq_data_merged_29-RNA_CGR_29",  # B.Subtilis day 14 aphids
  "RNAseq_data_merged_38-RNA_CGR_38", "RNAseq_data_merged_39-RNA_CGR_39", "RNAseq_data_merged_40-RNA_CGR_40", "RNAseq_data_merged_41-RNA_CGR_41",  # B.Subtilis day 1 no aphids
  "RNAseq_data_merged_47-RNA_CGR_47", "RNAseq_data_merged_48-RNA_CGR_48", "RNAseq_data_merged_49-RNA_CGR_49"                             # B.Subtilis day 14 no aphids
)

```

# LRT Testing

1.  Testing for time effect across A.radicis innoculated barley

```{r}
dds_subset <- dds[, dds$condition %in% c("Acidovorax_day1_aphids", "Acidovorax_day7_aphids", "Acidovorax_day14_aphids")]
dds_subset$time <- factor(gsub("Acidovorax_day|_aphids", "", dds_subset$condition))  # extract 1, 7, 14

design(dds_subset) <- ~ time + condition
dds_subset <- DESeq(dds_subset, test = "LRT", reduced = ~time)
res_time <- results(dds_subset)

head(res_time)
head(res_time[order(res_time$padj), ])

# Convert to dataframe
res_df <- as.data.frame(res_time)

# Add gene names as a column
res_df$gene <- rownames(res_df)

# Order by adjusted p-value
res_ordered <- res_df[order(res_df$padj), ]

# View top 10 genes
head(res_ordered, 10)

# Select and rename columns for clarity
res_table <- res_ordered[, c("gene", "baseMean", "stat", "pvalue", "padj")]
colnames(res_table) <- c("Gene", "BaseMean", "LRT_statistic", "Raw_pvalue", "Adjusted_pvalue")

# Preview nicely
print(res_table[1:10, ], row.names = FALSE)

write.table(res_table, 
            file = "LRT_acidovorax_timecourse_top_genes.tsv", 
            sep = "\t", 
            quote = FALSE, 
            row.names = FALSE)

View(res_table)

summary(res_time)

```

2.  Testing for time effect no bacteria + aphids

```{r}
# Subset for No Bacteria with aphids over days 1, 7, and 14
dds_subset_nb <- dds[, dds$condition %in% c("NoBacteria_day1_aphids", "NoBacteria_day7_aphids", "NoBacteria_day14_aphids")]

# Create time variable from condition
dds_subset_nb$time <- factor(gsub("NoBacteria_day|_aphids", "", dds_subset_nb$condition))  # extracts 1, 7, 14

# Set up the design and run LRT
design(dds_subset_nb) <- ~ time
dds_subset_nb <- DESeq(dds_subset_nb, test = "LRT", reduced = ~1)

# Get results
res_time_nb <- results(dds_subset_nb)

# Convert to dataframe and add gene names
res_df_nb <- as.data.frame(res_time_nb)
res_df_nb$gene <- rownames(res_df_nb)

# Order by adjusted p-value
res_ordered_nb <- res_df_nb[order(res_df_nb$padj), ]

summary(res_time_nb)
# Select key columns and rename
res_table_nb <- res_ordered_nb[, c("gene", "baseMean", "stat", "pvalue", "padj")]
colnames(res_table_nb) <- c("Gene", "BaseMean", "LRT_statistic", "Raw_pvalue", "Adjusted_pvalue")

# Preview top results
print(res_table_nb[1:10, ], row.names = FALSE)

# Save to a tab-separated file
write.table(res_table_nb, 
            file = "LRT_nobacteria_aphids_timecourse_top_genes.tsv", 
            sep = "\t", 
            quote = FALSE, 
            row.names = FALSE)

# Optional: view the table in RStudio
View(res_table_nb)

```

Bar chart of DEGS across days: A. radicis v no bacterias (aphids)

```{r}
# Retrieved DESeq2 results for Day 1 Acidovorax vs NoBacteria (both with aphids)
res_day1_acido_vs_nobac <- results(dds, contrast = c("condition", "Acidovorax_day1_aphids", "NoBacteria_day1_aphids"))

# Removed NAs from padj and filtered the results
res_filtered <- res_day1_acido_vs_nobac[!is.na(res_day1_acido_vs_nobac$padj), ]

# Converted the results to a dataframe
res_df <- as.data.frame(res_filtered)
res_df$gene <- rownames(res_df)

# Classified genes based on regulation status
res_df$regulation <- "Not Sig"
res_df$regulation[res_df$padj < 0.05 & res_df$log2FoldChange > 1] <- "Up"
res_df$regulation[res_df$padj < 0.05 & res_df$log2FoldChange < -1] <- "Down"

# Counted the number of DEGs per category
table(res_df$regulation)

# Filtered to include only Up/Down-regulated genes
res_day1_filtered <- res_df[res_df$regulation %in% c("Up", "Down"), ]

# Ensured "Up" and "Down" categories were included, even if the count was 0
deg_counts <- as.data.frame(table(factor(res_df$regulation, levels = c("Up", "Down"))))
colnames(deg_counts) <- c("Regulation", "Count")

# Plotted the bar chart for Day 1
ggplot(deg_counts, aes(x = Regulation, y = Count, fill = Regulation)) +
  geom_bar(stat = "identity", width = 0.6) +
  scale_fill_manual(values = c("Up" = "red", "Down" = "blue")) +
  labs(
    title = "Differentially Expressed Genes\nAcidovorax vs No Bacteria (Day 1, Aphids)",
    x = "Regulation Direction",
    y = "Number of Genes"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.title.x = element_text(face = "bold"),
    axis.title.y = element_text(face = "bold")
  )


# Retrieved differential expression results for Acidovorax vs NoBacteria (Day 7, Aphids)
res_day7 <- results(dds, contrast = c("condition", "Acidovorax_day7_aphids", "NoBacteria_day7_aphids"))

# Removed NAs and converted to dataframe
res_day7_df <- as.data.frame(res_day7[!is.na(res_day7$padj), ])
res_day7_df$gene <- rownames(res_day7_df)

# Classified genes as Up or Down based on thresholds
res_day7_df$regulation <- NA
res_day7_df$regulation[res_day7_df$padj < 0.05 & res_day7_df$log2FoldChange > 1] <- "Up"
res_day7_df$regulation[res_day7_df$padj < 0.05 & res_day7_df$log2FoldChange < -1] <- "Down"

# Filtered out "Not Sig" genes
res_day7_filtered <- res_day7_df[!is.na(res_day7_df$regulation), ]

# Ensured "Up" and "Down" categories were present and ordered correctly
deg_counts_day7 <- as.data.frame(table(factor(res_day7_df$regulation, levels = c("Up", "Down"))))
colnames(deg_counts_day7) <- c("Regulation", "Count")

# Plotted the Day 7 bar chart
ggplot(deg_counts_day7, aes(x = Regulation, y = Count, fill = Regulation)) +
  geom_bar(stat = "identity", width = 0.6) +
  scale_fill_manual(values = c("Up" = "red", "Down" = "blue")) +
  labs(
    title = "Differentially Expressed Genes\nAcidovorax vs No Bacteria (Day 7, Aphids)",
    x = "Regulation Direction",
    y = "Number of Genes"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.title.x = element_text(face = "bold"),
    axis.title.y = element_text(face = "bold")
  )

```

Generation of bar chart for DEGs: A.radicis v no bacteria (final)

```{r}

# Day 1 DEGs Processed
res_day1 <- results(dds, contrast = c("condition", "Acidovorax_day1_aphids", "NoBacteria_day1_aphids"))
res_day1_df <- as.data.frame(res_day1[!is.na(res_day1$padj), ]) #NA values removed
res_day1_df$gene <- rownames(res_day1_df)
res_day1_df$regulation <- "Not Sig"
res_day1_df$regulation[res_day1_df$padj < 0.05 & res_day1_df$log2FoldChange > 1] <- "Up"
res_day1_df$regulation[res_day1_df$padj < 0.05 & res_day1_df$log2FoldChange < -1] <- "Down"
res_day1_df$time <- "Day 1"

#Day 7 DEGs Processed
res_day7 <- results(dds, contrast = c("condition", "Acidovorax_day7_aphids", "NoBacteria_day7_aphids"))
res_day7_df <- as.data.frame(res_day7[!is.na(res_day7$padj), ]) #NA values removed
res_day7_df$gene <- rownames(res_day7_df)
res_day7_df$regulation <- "Not Sig"  # FIXED: use "Not Sig" for consistency
res_day7_df$regulation[res_day7_df$padj < 0.05 & res_day7_df$log2FoldChange > 1] <- "Up"
res_day7_df$regulation[res_day7_df$padj < 0.05 & res_day7_df$log2FoldChange < -1] <- "Down"
res_day7_df$time <- "Day 7"

#Day 14 DEGs Processed
res_day14 <- results(dds, contrast = c("condition", "Acidovorax_day14_aphids", "NoBacteria_day14_aphids"))
res_day14_df <- as.data.frame(res_day14[!is.na(res_day14$padj), ])#NA values removed
res_day14_df$gene <- rownames(res_day14_df)
res_day14_df$regulation <- "Not Sig"
res_day14_df$regulation[res_day14_df$padj < 0.05 & res_day14_df$log2FoldChange > 1] <- "Up"
res_day14_df$regulation[res_day14_df$padj < 0.05 & res_day14_df$log2FoldChange < -1] <- "Down"
res_day14_df$time <- "Day 14"

#All Days combined
all_data <- bind_rows(res_day1_df, res_day7_df, res_day14_df)

#Counting of Up/Down Regulated Genes
deg_counts_all <- all_data %>%
  filter(regulation %in% c("Up", "Down")) %>%
  count(time, regulation, name = "Count")

# time/regulation combination appear even with 0 counts (day 14)
deg_counts_all_complete <- expand.grid(
  time = c("Day 1", "Day 7", "Day 14"),
  regulation = c("Up", "Down")
) %>%
  left_join(deg_counts_all, by = c("time", "regulation")) %>%
  replace_na(list(Count = 0)) 

# Set time order explicitly
deg_counts_all_complete$time <- factor(deg_counts_all_complete$time, levels = c("Day 1", "Day 7", "Day 14"))

# Plotting of 3 panel bar chart (day 1,7, 14)
ggplot(deg_counts_all_complete, aes(x = regulation, y = Count, fill = regulation)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.6) +
  scale_fill_manual(values = c("Up" = "red", "Down" = "blue")) +
  labs(
    title = "Differentially Expressed Genes\nAcidovorax vs No Bacteria (Aphids)",
    x = "Regulation Direction",
    y = "Number of DEGS"
  ) +
  facet_wrap(~ time, scales = "free_y") +
  scale_y_continuous(limits = c(0, 60), breaks = seq(0, 60, 5)) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.title.x = element_text(face = "bold"),
    axis.title.y = element_text(face = "bold")
  )



```

DEGs across days B.subtilis inoculated mapped to A.radicis genome

```{r}
# Day 1 DEGs: B. subtilis vs No Bacteria (Aphids)
res_day1 <- results(dds, contrast = c("condition", "Bsubtilis_day1_aphids", "NoBacteria_day1_aphids"))
res_day1_df <- as.data.frame(res_day1[!is.na(res_day1$padj), ])
res_day1_df$gene <- rownames(res_day1_df)
res_day1_df$regulation <- "Not Sig"
res_day1_df$regulation[res_day1_df$padj < 0.05 & res_day1_df$log2FoldChange > 1] <- "Up"
res_day1_df$regulation[res_day1_df$padj < 0.05 & res_day1_df$log2FoldChange < -1] <- "Down"
res_day1_df$time <- "Day 1"

# Day 7 DEGs
res_day7 <- results(dds, contrast = c("condition", "Bsubtilis_day7_aphids", "NoBacteria_day7_aphids"))
res_day7_df <- as.data.frame(res_day7[!is.na(res_day7$padj), ])
res_day7_df$gene <- rownames(res_day7_df)
res_day7_df$regulation <- "Not Sig"
res_day7_df$regulation[res_day7_df$padj < 0.05 & res_day7_df$log2FoldChange > 1] <- "Up"
res_day7_df$regulation[res_day7_df$padj < 0.05 & res_day7_df$log2FoldChange < -1] <- "Down"
res_day7_df$time <- "Day 7"

# Day 14 DEGs
res_day14 <- results(dds, contrast = c("condition", "Bsubtilis_day14_aphids", "NoBacteria_day14_aphids"))
res_day14_df <- as.data.frame(res_day14[!is.na(res_day14$padj), ])
res_day14_df$gene <- rownames(res_day14_df)
res_day14_df$regulation <- "Not Sig"
res_day14_df$regulation[res_day14_df$padj < 0.05 & res_day14_df$log2FoldChange > 1] <- "Up"
res_day14_df$regulation[res_day14_df$padj < 0.05 & res_day14_df$log2FoldChange < -1] <- "Down"
res_day14_df$time <- "Day 14"

# Combine all time points
all_data <- bind_rows(res_day1_df, res_day7_df, res_day14_df)

# Count Up/Down regulated genes
deg_counts_all <- all_data %>%
  filter(regulation %in% c("Up", "Down")) %>%
  count(time, regulation, name = "Count")

# Ensure all combinations are present
deg_counts_all_complete <- expand.grid(
  time = c("Day 1", "Day 7", "Day 14"),
  regulation = c("Up", "Down")
) %>%
  left_join(deg_counts_all, by = c("time", "regulation")) %>%
  replace_na(list(Count = 0)) 

# Set time order
deg_counts_all_complete$time <- factor(deg_counts_all_complete$time, levels = c("Day 1", "Day 7", "Day 14"))

# Plot
ggplot(deg_counts_all_complete, aes(x = regulation, y = Count, fill = regulation)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.6) +
  scale_fill_manual(values = c("Up" = "red", "Down" = "blue")) +
  labs(
    title = "Differentially Expressed Genes\nB. subtilis vs No Bacteria (Aphids)",
    x = "Regulation Direction",
    y = "Number of DEGs"
  ) +
  facet_wrap(~ time, scales = "free_y") +
  scale_y_continuous(limits = c(0, 60), breaks = seq(0, 60, 5)) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.title.x = element_text(face = "bold"),
    axis.title.y = element_text(face = "bold")
  )


```

```{r}
# DE results for pairwise comparison
res_day1_bsub_vs_noaphid <- results(dds, contrast = c("condition", "Bsubtilis_day1_aphids", "Bsubtilis_day1_no_aphids"))

# Removal of NA padj values
res_filtered <- res_day1_bsub_vs_noaphid[!is.na(res_day1_bsub_vs_noaphid$padj), ]

# Convert to dataframe & add gene names
res_df <- as.data.frame(res_filtered)
res_df$gene <- rownames(res_df)

# Add regulation category based on log fold change
res_df$regulation <- "Not Sig"
res_df$regulation[res_df$padj < 0.05 & res_df$log2FoldChange > 1] <- "Up"
res_df$regulation[res_df$padj < 0.05 & res_df$log2FoldChange < -1] <- "Down"

# Volcano plot
ggplot(res_df, aes(x = log2FoldChange, y = -log10(padj), color = regulation)) +
  geom_point(alpha = 0.6) +
  scale_color_manual(
    values = c("Up" = "red", "Down" = "blue", "Not Sig" = "grey"),
    labels = c("Up" = "Overexpressed", "Down" = "Underexpressed", "Not Sig" = "Not Significant")
  ) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black") +
  geom_vline(xintercept = 0, linetype = "dotted", color = "black") +
  labs(
    title = "Volcano Plot: B. subtilis Day 1 (Aphids) vs B. subtilis Day 1 (No Aphids)",
    x = "log2 Fold Change",
    y = "-log10 adjusted p-value"
  ) +
  guides(color = guide_legend(title = "Expression")) +
  theme_minimal()


```

Volcano plots B.subtilis v no bacteria day 7 (aphids)

```{r}
# DE results for Day 7 aphid comparison
res_day7_acido_vs_nobac_aphids <- results(dds, contrast = c("condition", "Bsubtilis_day7_aphids", "NoBacteria_day7_aphids"))

# Removal of NA padj values
res_filtered <- res_day7_acido_vs_nobac_aphids[!is.na(res_day7_acido_vs_nobac_aphids$padj), ]

# Converted to dataframe & add gene names added
res_df <- as.data.frame(res_filtered)
res_df$gene <- rownames(res_df)

# Regulation category added for up or downregulated genes based on log fold change
res_df$regulation <- "Not Sig"
res_df$regulation[res_df$padj < 0.05 & res_df$log2FoldChange > 1] <- "Up"
res_df$regulation[res_df$padj < 0.05 & res_df$log2FoldChange < -1] <- "Down"

# Filtering of significant genes
top_genes <- res_df[order(res_df$padj), ][1:10, ]

# Load ggplot2 and ggrepel
library(ggplot2)
library(ggrepel)

# Volcano plot generated with top 10 significant genes labeled (without label lines)
ggplot(res_df, aes(x = log2FoldChange, y = -log10(padj), color = regulation)) +
  geom_point(alpha = 0.6) +
  scale_color_manual(
    values = c("Up" = "red", "Down" = "blue", "Not Sig" = "grey"),
    labels = c("Up" = "Overexpressed", "Down" = "Underexpressed", "Not Sig" = "Not Significant")
  ) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black") +  # Significance threshold line
  geom_vline(xintercept = 0, linetype = "dotted", color = "black") +             # log2FC = 0 reference
  geom_text_repel(
    data = top_genes,
    aes(label = gene),
    size = 3,
    box.padding = 0.3,
    point.padding = 0.3,
    max.overlaps = Inf,
    segment.color = NA  # Remove connecting lines from labels
  ) +
  labs(
    title = "Volcano Plot: Acidovorax vs No Bacteria (Day 7, With Aphids)",
    x = "log2 Fold Change",
    y = "-log10 adjusted p-value"
  ) +
  guides(color = guide_legend(title = "Expression")) +
  theme_minimal()
```
